---
title: "The Effects of Partial Sleep Deprivation on Mind Wandering, Behaviour and Mood"
shorttitle: "PSD on MW, EFs and Mood"
authors:
  - name: Steffen Rygg Aasen
    orcid: 0000-0003-2807-7504
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - writing
      - data curation
      - investigation
      - formal analysis
      - resources
      - visualization
  - name: Samy Engebø Babiker
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - data curation
      - investigation
      - resources
      - editing
  - name: Ingvild Saksvik-Lehouillier
    orcid: 0000-0002-1953-7364
    affiliation: Norwegian University of Science and Technology
    roles: 
      - resources
      - methodology
      # - editing
  # - name: Julia W. Y. Kam
  #   orcid: 0000-0002-2369-2148
  #   affiliation: Helen Wills Neuroscience Institute, University of California – Berkeley, Berkeley, CA, USA
  #   roles: 
  #     - editing
  - name: Aurora Marie Kleivik Vangen
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - data curation
      - investigation
      - resources
  - name: Silje Balsnes Haugan
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - data curation
      - investigation
      - resources
  - name: Sanna Aurora Sandell
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - data curation
      - investigation
      - resources
  - name: Gábor Csifcsák
    orcid: 0000-0001-7719-7073
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - conceptualization
      - methodology
      - supervision
      - validation
      - resources
      - editing
  - name: Matthias Mittner
    orcid: 0000-0003-0205-7353
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - conceptualization
      - funding acquisition
      - project administration
      - methodology
      - software
      - resources
      - supervision
      - validation
      - visualization
      - editing
    corresponding: true
abstract: "Insufficient sleep is a common occurrence in modern society and has several consequences on our attention, executive functioning (EF), and mood. Yet, there is lacking research on the effects of partial sleep deprivation (PSD) on the interplay between them. Moreover, attentional states are often not convoluted but rarely dissected. To elucidate on these issues, we had 34 healthy adults engage in a PSD protocol, where they slept on average 1.7 hours less for 3 consecutive nights before a test, which were compared against themselves after a period of normal sleep (NS, counter-balanced). Participants engaged in the finger-tapping random sequence generation task (FT-RSGT) that provided subjective reports of their attentional state (on-task or mind wandering, MW) and whether they had any thought content (or were blanking), as well as the intention of their attentional state. The task also provided objective measures of performance as variability in the responses and the randomness of the sequences both of which have been related to attentional states. The reduction in sleep resulted in increased MW, mind blanking, response variability, negative mood, and decreased positive mood, as well as reduced randomness over time. These findings highlight the importance of sufficient sleep in maintaining sustained attention, optimal task performance and good mood."
keywords: [attention, sleep, mind wandering, partial sleep deprivation, executive functions, mood]
author-note:
  disclosures:
    conflict-of-interest: "The author has no conflict of interest to declare."
    data-sharing: "Data from this study can be accessed at the Open Science Foundation (OSF): https://osf.io/xq6wr/."
    financial-support: "WHAT ARE WE SUPPORTED BY? "
    gratitude: "We would like to extend our gratitude to Ragnhild Nicolaisen Drevland for helping in the data collection. Moreover, we would like to express our deep gratitude to all the participants slept less (for Science!)."
bibliography: references.bib

# Format
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf: default
  apaquarto-typst: default

# Code settings
editor: source
execute:
  include: false
  echo: false
---

```{r load project data munge Parameters, warning=FALSE}
relative_path <- "../" # b/c quarto 
source( paste0(relative_path, "lib/quarto_load_project.R") )
quarto_load_project(relative_path, load_data_data = T)
```

# Introduction

Sufficient sleep is paramount for optimal functioning in our everyday lives [@alhola2007] and insufficient sleep demonstrates far-reaching consequences across many different health outcomes, such as depression and obesity [@chattu2019]. Moreover, numerous studies demonstrate the negative influence of sleep restriction on sustained attention, executive functions, emotional stability, and memory consolidation [@åkerstedt2007; @alhola2007; @anderson2011; @lo2016; @poh2016; @stickgold2007]. For example, sleep deprivation increases attentional lapses and impulsivity during sustained attention tasks [@anderson2011; @lo2016; @saksvik-lehouillier2020], and decreases speed of processing during an arithmetic task [@lo2016]. Insufficient sleep also impedes memory consolidation, likely, by disrupting neural plasticity processes that occur during sleep, especially those associated with slow-wave sleep and rapid-eye-movement sleep [@stickgold2007] {{{I suppose talking about memory consolidation is a bit beside out research here, and probably best left aside to not distract the reader? }}}. Additionally, sleep loss can reduce positive [@lo2016; @saksvik-lehouillier2020] and increase negative emotions [@cárdenas-egúsquiza2022b] influencing our overall well-being in a way that might predispose individuals towards depression [@chaieb2022].

Most studies implement total sleep deprivation (TSD) where participants who are not allowed to sleep at all are contrasted with those following their regular sleep schedule. However, TSD is less common compared to partial sleep deprivation [PSD; e.g., @sheehan2019; @sivertsen2019], where individuals sleep less than the recommended duration (e.g., an adult sleeping 5 hours instead of the recommended 7-8 hours), an effect that can accumulate over multiple days. Intriguingly, the accumulating effect of PSD might impact cognitive functions and mood more than TSD [@pilcher1996; @salfi2020]. In general, PSD may be less noticeable than TSD because people can habituate to the reduced amount of sleep [@balkin2008] and develop coping strategies to compensate for the reduced cognitive functioning [@rupp2009]. Moreover, PSD appears to require multiple days of normal sleep (NS) to recover normal cognitive functioning whereas a single night is often sufficient after TSD \[@balkin2008; @belenky2003; @lo2016; @cohen2010\]. In summary, despite the prevalence of PSD [e.g., @sheehan2019; @sivertsen2019], its effects remain surprisingly underexplored [@alhola2007].

Insufficient sleep influences most cognitive functions, including attention. Sleep deprivation worsen task focus by increasing shifts of attention away from the task to internal thoughts, a phenomenon researchers usually call mind wandering [MW; @smallwood2015]. For example, one study indicated that one night of TSD increased MW [@poh2016]. Additionally, recent evidence suggests that the effect between MW and sleep might be directional. Two diary studies found that self-reported sleep disturbances predicted increased MW the following day [@cárdenas-egúsquiza2022b], while more MW during the day did not predict sleep problems the following night [@marcusson-clavertz2019; but see @fell2024].

Given the causal evidence showing that sleep deprivation increases MW, MW might be one of the underlying consequences of reduced cognitive performance following sleep deprivation. Indeed, similar behavioural consequences are often observed following sleep deprivation and MW episodes. For example, in a simulated driving environment, participants show greater driving speed variation, lane variation, and braking response times following sleep deprivation [@jackson2013]. Similarly, MW during a simulated driving environment increase driving speed and braking response times [@rodd2017; @yanko2014]. Furthermore, during more formal cognitive tests, sleep deprivation has been found to increase response variability [@jackson2013] and reduce random sequence generation [@heuer2005], similar to how MW episodes are related to increased response variability [@boayue2020; @aasen2024; @alexandersen2022; @baldwin2017; @seli2013b] and reduced random sequence generation [e.g., @drevland2024; @teasdale1995].

The tendency to shift our attention is multifaceted and complex. For example, MW episodes typically have a specific reportable content, like thinking about what to make for dinner or upcoming social events. Occasionally, however, we might find ourselves not focusing on the task but at the same time not thinking about anything. This lack of content during inattention is referred to as mind blanking (MB), and proposals have suggested that MB as well as MW might be increased by sleep deprivation [@andrillon2019; @jubera-garcía2021sleepMW] - something that was recently confirmed [@boulakis2024]. Furthermore, research has demonstrated that MB is associated with slower response times, increased misses [@andrillon2021; @mortaheb2022], reduced pupil size, and increased subjective sleepiness [@unsworth2018] compared to MW. These associations suggest that MB may be linked to reduced information processing [@boulakis2023; @musat2024; @mortaheb2022] plausibly through deeper task disengagement [i.e., stronger perceptual coupling; @smallwood2015; e.g., @ward2013] and lower arousal [@boulakis2024]. Altogether, these findings indicate that MB can be distinguished from MW, and that both are influenced by sleep deprivation.

The tendency to shift our attention can be done deliberately or happen spontaneously. Research suggests that participants with lower motivation report more MW and are engaged in more deliberate (D-MW) than spontaneous MW [S-MW; @robison2018; @seli2015]. Moreover, participants seem to engage more in S-MW than D-MW during difficult tasks, and, conversely, more D-MW than S-MW during easy tasks [@seli2016]. Additionally, research has demonstrated a negative relationship between working memory capacity (WMC) and S-MW [@ju2018; @robison2018] suggesting that WMC plays a role in regulating MW intention. Given the negative impact of sleep deprivation on both executive functioning [EF; @heuer2005] and motivation [@engle-friedman2003; @killgore2010], it is plausible that sleep-deprived individuals may experience alterations in both their capacity and intention to engage in different forms of MW.

Variation in sleep has also been related to variations in mood. For example, sleep-deprived individuals tend to reported less positive and more negative mood, while also indicating more anxiety and depressive symptoms [@palmer2024; @short2020]. Similarly, MW has also been associated with negative affect in the general population [@killingsworth2010], and is increased in patients suffering from various psychiatric disorders, such as depression and anxiety disorder [@marchetti2012; @marchetti2016; @smallwood2007b]. Moreover, it has been proposed that positive mood might be related to motivation [@erez2002], which has been shown to be reduced after sleep deprivation [@engle-friedman2003] and can influence MW tendency [@seli2015]. Thus, it is possible that the changes in mood following sleep deprivation could influence the tendency to mind wander and its various manifestations.

As noted, EF is negatively influenced by sleep deprivation [@heuer2005], yet much research on MW does not adequately load executive functions [@boayue2020]. The finger tapping random sequence generation task [FT-RSGT; @boayue2020], however, effectively measures EF while also capturing MW tendencies. The FT-RSGT requires participants to create a random sequence of responses using two dedicated keys, while matching the rhythm of an ongoing metronome as closely as possible. Producing random sequences engages the central executive [@baddeley1998], by requiring inhibitory control of pre-potent responses, as well as mental monitoring and updating of previously generated sequences, all of which are core functions of the executive system [@miyake2000]. The randomness of the sequences has been negatively related to MW [e.g., @teasdale1995; @boayue2020; @aasen2024; @groot2022], while response variability has been positively associated with MW [@boayue2020; @alexandersen2024; @bastian2013; @cheyne2009; @seli2013b].

To unravel the complex relationship between MW (and its various manifestations), the executive system, mood, and sleep, we conducted a within-subject, cross-over study using a PSD protocol [@saksvik-lehouillier2020]. Our main hypotheses concerned the effect of PSD on MW and FT-RSGT performance. Specifically, we expected that PSD would increase participants’ overall propensity to engage in MW, increase the frequency of MB episodes relative to MW with reportable content, lead to more S-MW relative to D-MW, increase negative mood, reduce positive mood, increase BV and reduce randomness generation in the FT-RSGT. Additionally, we expected that PSD-associated reduction in mood before the task would predict the amount of MW during the task. Moreover, we also aimed to assess how PSD would influence the interaction between measures of EF and different forms of MW, and whether PSD and changed mood scores would moderate the relationship between task-focus and EF performance.

# Method

## Participants

We recruited `r subj$full$num` healthy adults with an average age of `r subj$full$age` (*SD* = `r subj$full$age_sd`, range = `r subj$full$age_min`-`r subj$full$age_max`, females = `r subj$full$gender_f`) to a within-subject cross-over design study {{{This is noted in the intro}}}. In the analyses reported here, no participants were excluded. In the Supplementary, however, we report analyses for a reduced sample (see below) where `r subj$red$exc` participants were excluded, leaving us with `r subj$red$num` participants with an average age of `r subj$red$age` (*SD* = `r subj$red$age_sd`, range = `r subj$red$age_min`-`r subj$red$age_max`, females = `r subj$red$gender_f`). Data-collection took place between March 2021 and May 2022.

The experiment was approved by the institutional ethics board at UiT The Arctic University of Norway. Participants were recruited via social media posts, flyers, and personal contacts in Tromsø. To be included, participants needed to be between 18 and 50 years, with no history of or ongoing mental illness, as well as not taking any psychoactive drugs. Participants were also asked whether they had any hearing impairments (used to adjust the metronome sound, but none reported any impairments). Lastly, participants had to give written consent to participate, and they were compensated for their time with a gift card at a local shopping mall valued at 500 NOK (approximately 50 USD).

## The Task

The task (@fig-protocol-a) was coded in PsychoPy, Standalone version 1.83.04_win32 [@peirce2007]. The metronome had an inter-stimulus interval (ISI) of 750 ms where it played a sound of 440 Hz for 75 ms on two stereo speakers placed beside the monitor, approximately 0.5 meter from the participant, with a peak volume of approximately 60 dB. Furthermore, participants were pseudo-randomly probed about their attentional state. The probes were presented randomly every 40 to 80 seconds. Each time, participants were asked three consecutive questions relating to their attention, which were answered on a 4-point Likert scale. The first question was related to their focus: “To what degree were you focused on the task just before this question?” (1: “clearly unfocused”, 4: “clearly focused”). The second question was related to the content of the MW episode: “To the degree to which you were not focusing on the task, were you thinking of something in particular or just thinking about nothing?” (1: “clearly nothing”, 4: “clearly something specific”). Lastly, we wanted to distinguish between the intentionality of the MW episode: “Were you deliberate about where you focused your attention (either on-task or elsewhere) or did it happen spontaneously?” (1: “clearly spontaneous” to “clearly deliberate”). The task lasted for approximately 25 minutes answering a total of 25 probes. All material in this study is publicly available in Norwegian at the Open Science Framework (OSF; <https://osf.io/xq6wr/>; partial English translations can be found at <https://osf.io/cv24f/> and <https://osf.io/txpu2/>).

Both BV and AE were estimated from the 25 stimuli preceding a thought probe, corresponding to 18.75 seconds [@boayue2020]. BV was calculated as the log-transformed standard deviation of the inter-tap-intervals and then z-scored across all subjects. The AE variable [@pincus1997] was parametrized by parameter *m* that indicates the maximum length of subsequence’s used for the entropy-calculation. For the current analysis, we used *m* = 2, which has been found to provide the most stable estimates with a robust effect in relation to MW [@boayue2020]. Raw AE values were transformed using $-log(log(2)-AE)$ and then z-scored across all subjects. Higher values of AE imply greater irregularities (more random responding) of the sequence [@pincus1991], associated with stronger recruitment of executive resources [@teasdale1995].
::: {#fig-protocol layout-nrow="2" fig-cap-location="top"}
![Sleep protocol](../outputs/experiment/sleep-dot.svg){width="50%"}

![Finger-tapping random sequence generation task](../outputs/experiment/FT-RSGT_overview.svg){width="60%"}
:::

*Note*. (Top) At baseline (T0) participants signed the consent, received the questionnaires, and the actigraphy device, and were scheduled for their test days. At the first test (T1) participants performed the finger-tapping random sequence generation task (FT-RSGT), and data from the actigraphy was retrieved. The second test (T2) followed the same procedure as T1 (except for collecting the actigraphy and the diary). Both groups followed the normal sleep (NS) protocol for the first 7 nights but diverged afterwards. The late sleep deprivation (LSD) group took their first test (T1) immediately after the 7 NS nights, while the early sleep deprivation (ESD) immediately proceeded to the partial sleep deprivation (PSD) nights. Since we wanted participants to sleep normally for 7 nights prior to their NS test, the ESD group ended up with three additional days compared to the LSD group. The dark-red dots represent meetings in the lab (T0 = first, T1 = second and T2 = third meeting), black dots represent non-study days between the first lab meeting (T0) and the onset of data collection represented by the light blue dots that also represents the normal sleep (NS) nights, and the purple dots represents the partial sleep deprivation (PSD) nights. The red (downward) arrows indicate experimental retrieval of data (e.g., consent as a pen). The green (upward) arrows indicate material given to the participants, and the orange circle arrow indicate retrieval of data but return of the device (i.e., actigraphy). (Bottom) An example of how behavioural variability (BV) and approximate entropy (AE) change over time in the finger-tapping random sequence generation task (FT-RSGT). BV is estimated as the inter-tap interval (ITI) between the taps. Higher values indicate greater variability, i.e., worse performance, while lower values indicate less variability, i.e., better performance - illustrated via the grey line. AE is related to the response type (blue/red). Higher AE indicate more random sequences, while lower AE indicate less random sequences (solid black line).

## Sleep Paradigm

<!--# Paradigm in general  -->

To investigate the effect of PSD on MW and executive functions, we used a modified PSD protocol described by [@saksvik-lehouillier2020; @fig-protocol-b]. In their study, the authors asked their participants to sleep 2 hours less than usual each night for 3 consecutive nights, and their participants achieved an average of 2.23 (*SD* = 0.72) hours less sleep according to their actigraphy measure. To increase the effectiveness of the PSD manipulation [@lim2008; @saksvik-lehouillier2020; @schwarz2016], we asked our participants to sleep 3 hours less for three consecutive nights in the PSD condition. During normal sleep, participants were instructed to not go to sleep later than 00:30 and not wake up later than 09:00. Meanwhile, during the PSD, they were instructed to go to sleep 3 hours later than usual, but still wake up at the same time as during their normal sleep

<!--# Setup of the protocol -->

The order of the conditions was counter-balanced across subjects, resulting in two groups: Participants in the early sleep deprivation group (ESD) underwent PSD before the first data-collection, while participants in the late sleep deprivation group (LSD) had a period of controlled NS before the first data collection (see @protocol-fig). Participants were pseudo-randomly assigned to either the LSD or ESD condition, but we were flexible with group assignment to better fit the participants' personal schedule. Data from both groups were merged for the analysis.

!["Sleep protocol"](../outputs/experiment/sleep_dot.svg){#fig-protocol-b width="50%"}

<!--# Measuring the compliance -->

To ensure compliance with the sleep protocol, we used two measures of participants' sleep. Firstly, participants were required to wear an Actiwatch Spectrum Pro (Philips Respironics, USA; <https://tinyurl.com/yjr2szxg>) during their sleep, but were recommended to wear it all the time. The Actiwatch measures physical activity and ambient light which is used to estimate participants' sleep- and waking-time, sleep latency and total sleep using the accompanying software, Philips Actiware 6.0.2. Secondly, participants filled out a sleep diary in the morning after waking. They were asked when they went to bed, when they tried to fall asleep, how long it took before they fell asleep, how many times they woke up during the night before waking up in the morning, how long these wake-ups were, when the last wake-up occurred, when they rose from the bed, to rank their sleep quality from very bad to very good (5-point Likert scale), and if they took off the Actiwatch at any time the day before. To calculate the participants’ sleep duration, we first estimated their sleep onset by adding their self-reported bedtime and their estimated sleep delay. Participants' wake times were defined as their self-reported last waking time, and then, their sleep duration was calculated by taking the difference between their sleep onset and their wake time. For all sleep estimates, we used the 3 nights preceding each test because the PSD condition was limited to 3 nights of sleep restriction.

<!--# Discrepancies between the measures -->

The self-reported and actigraphy estimated sleep onset and wake times did not always converge (see @fig-SD-between-modalities). Cases were considered diverging when they differed by more than 15 minutes. To resolve these cases, we loosely followed the procedure described by @follesø2021 to estimate an adjusted sleep duration. The procedure provides a standardized, hierarchical method to redefine rest intervals before falling asleep and after waking up in the morning, by systematically comparing algorithm-generated estimates with event markers, sleep diary entries, and changes in activity and light patterns.

<!--# The Shiny application -->

To re-evaluate the divergent cases, we created a Shiny application. The application visualized time on the x-axis and the actigraphy recorded activity (raw) and light (log transformed) on separately y-axis. Coloured marks highlighted the self-reported and actigraphy sleep onset and wake time. The rest interval was indicated through shaded areas ranging from the sleep attempt to the sleep onset, and the wake time to the rise time for both modalities. Two sliders were provided (hours and minutes) to indicate the new onset/wake time, but separate buttons were provided to apply the exact values of the self-reported or actigraphy onset/wake times.

<!--# Resolving the discrepancies -->

We had two researchers (S.R.A. and G.C.) individually reevaluate the sleep in general terms with the procedure by @follesø2021. After the evaluation, there was a high level of agreement between the researchers (wake *r* = `r sleep_adjustment_round1_df |> filter(cond=="wake") |> pull(cor) |> fmt_APA_numbers(.low_val = T)`, onset *r* = `r sleep_adjustment_round1_df |> filter(cond=="onset") |> pull(cor) |> fmt_APA_numbers(.low_val = T)`). However, to reach full agreement, the researchers collectively evaluated those cases diverging by more than 15 minutes of those cases they had evaluated, while cases below 15 minutes were averaged. Importantly, since we incorporated both the actigraphy and self-report, we relied on either type (e.g., self-report) if a report from the other (e.g., actigraphy) was missing. Thus, the adjusted sleep benefited from a more complete dataset for the sleep estimation. For our analyses, we chose to use the adjusted sleep estimates, as it resolved some obvious discrepant cases. Importantly, we treated the PSD variable as a *continuous*, rather than as a dichotomous variable (condition), relating to the relative sleep difference between the NS and the PSD condition for each subject respectively.

```{r FIG - sleep deprivation between source type}
#| label: fig-SD-between-modalities
#| fig-cap: Difference in Sleep Duration Between Sources 
#| fig-cap-location: top
#| apa-note: "Sleep deprivation compared to the normal sleep (NS). Sleep duration is averaged for the three nights preceding each test, and the difference is caluclated as partial sleep deprivation (PSD) - NS. Black solid line indicate the original acceptance threshold, while the red dashed line indicated the adjusted sleep threshold (the criterion used for the reduced sample). Cases above the 1.5 threshold are highlighted by their associated participant code."
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-format: svg
#| fig-width: 10

outputs[["figs"]][["PSD__report_type"]]
```

# Procedure

### Baseline

At the start of the study (T0) participants were informed about the experimental protocol and had to give written consent before proceeding. Once written consent was given, participants were handed out a Phillips Actiwatch, a sleep diary, and a general questionnaire about their sleep patterns. In the general sleep questionnaire, we asked whether the participants had any night shifts, if they worked, with four discrete options: (1) nothing, (2) one night or less a month, (3) 2-4 nights per month or (4) 2 nights per week or more. Furthermore, we asked about the participants’ sleep preference during the weekdays and weekends: when they preferred to fall asleep, wake up, and for how long they preferred to sleep. They were then asked about how many hours they slept in a typical 24-hour period during the weekdays and weekends. Lastly, we inquired how many caffeine-containing beverages (e.g., coffee, tea) and energy drinks they usually consume in a normal day [@saksvik-lehouillier2020].

{{{MM: The below can be moved to the supplementary, and a shorter summary can be written, if space becomes an issue. }}}

We used verified scales to measure sleep, fatigue, insomnia, and sleep quality [@saksvik-lehouillier2020]. Participants' fatigue was measured with the Fatigue Severity Scale [@krupp1989] with lower scores indicating less fatigue. To measure sleepiness, we used the Epworth Sleepiness Scale [@johns1991], where lower scores (0-5) indicate low daytime sleepiness while higher scores (16-24) indicate severe excessive daytime sleepiness. Sleeping problems were measured with the Insomnia Severity Index [@bastien2001] where lower scores (0-7) indicate no significant insomnia, while higher scores (22-28) indicate severe insomnia. Participant’s sleep quality was measured with the Pittsburgh Sleep Quality Index [@buysse1989] that inquires about participants’ sleep time, sleep delay, wake time and sleep amount, where higher scores indicating worse sleep quality. Participant’s diurnal preferences were assessed with the Diurnal Scale [@torsvall1980] where higher scores indicate a preference to wake up and go to sleep early (‘early birds’), and lower scores indicate a preference to go to sleep later and wake up late (‘night owls’). To measure the participant’s mood, we used the Positive and Negative Affect Schedule [PANAS, @watson1988] where higher scores (min 10, max 50) indicate more intensive emotional states (for our analyses the PANAS was rescaled to range from 0 to 40). Lastly, we also asked participants about their alcohol consumption habit, derived from [@saksvik-lehouillier2020]. This consisted of 3 different questions with 5 answer alternatives (ranging from low to high consumption).

### Data Collection

Data collection procedures[^1] were identical for the LSD and ESD groups. LSD participants came to the first session after 7 days of NS, while in the ESD group, the session commenced after 3 additional days of PSD.

[^1]: Due to the SARS-CoV-2 pandemic in 2020-2022, we followed the safety guidelines given by our faculty. All surfaces touched by the participants were cleaned after each session, the experimenters wore disposable protective gear (e.g., gloves and facemasks), and participants were provided with a facemask if they did not already wear one. During testing, participants could take off their facemasks, but had to put it back on after the test was completed.

When participants arrived in the laboratory, they completed the PANAS to assess pre-session mood and were provided with the instructions for the FT-RSGT. During the experiment, we measured electroencephalography (EEG) activity using a 32 channel BrainAmp system (Brainproducts, Munich, Germany) with Ag/AgCl electrodes (data not analysed here). Participants were seated in an isolated sound-proofed compartment approximately 50 cm in front of a computer screen (Sony Trinitron CPD-E530, 21 inches) with two loudspeakers beside it. During the experiment, participants kept their head in a chin rest and were instructed to sit as still as possible and keep their gaze at the fixation cross on the screen. Before starting the task, participants engaged in a 1-minute-long training round of the FT-RSGT which could be repeated as many times as necessary. After the task, lasting approximately 25 minutes, participants again fill out a PANAS to measure their post-task mood. Then they filled out a custom-made questionnaire regarding the retrospective task motivation (7-point Likert scale ranging from “not very motivated” to “extremely motivated”), and to rate their confidence when responding to the first probe about their task focus (7-point Likert scale ranging from “not at all” to “extremely”). Furthermore, participants provided open-ended feedback about their randomness strategy, comments relating to the task, the content of their MW episodes, and other remarks regarding the experiment in general. If it was the participants first test (T1), they were reminded of their second test (T2) which was always scheduled seven days after their T1 at the same time (either 09:00 or 11:00). If it was their second test (T2), participants received their gift card and were asked whether they wanted a computer-generated summary of their sleep data.

## Statistical analysis

<!--# Bayesian analyses -->

The data analysis was done in `r str_split(R.version$version.string, " \\(")[[1]][[1]]` [@rcoreteam2025]. For our data analyses, we used Bayesian statistics [@wagenmakers2018] with the brms [Bayesian Regression Models using Stan; @bürkner2017] package. For our regression models, we used the default priors, which are non-informative for coefficients corresponding to fixed effects and weakly informative for the intercept and standard-deviation parameters (Student-t prior with 3 df, mean 0 and SD 2.5). The brms package relies on the Stan software [@standevelopmentteam2024] implementing Hamiltonian Mote Carlo sampling. For our models, we ran 6 chains with 6000 iterations each and a warm-up of 3000 iterations. Convergence and stability of the Bayesian sampling was assessed using $\hat{R}$, was ensured to be below 1.05 for all variables [@gelman2019]. We also ensured that the effective sample size (ESS) was greater than 1000 [@bürkner2017]. When reporting Bayesian regression coefficients, we report the posterior mean *b* and its 95% highest-density interval (HDI). In addition, we report the evidence ratio (ER) that the effect is in the (*b*) specified direction. The ER indicates the relative evidence for the effect being in the (*b*) specified ($\text{ER}_\text{dir}$) vs. the opposite direction. For example, a coefficient of 1.2 with an ER of 30 indicates that there is 30 times more evidence that the effect is *positive* (as specified by the *b*) rather than negative. Finally, we report the probability that the effect is in the (*b*) specified direction [$p_\text{dir}$; e.g., $b = -1.2$, $p_\text{dir} = .967$; indicates that there is a 96.7% probability that the effect is *negative* rather than positive; @boayue2020].

<!--# Models - Thought Probes -->

To analyse the thought probes, we implemented Bayesian hierarchical ordinal-probit models [@bürkner2019] to avoid problems with using standard linear models [@boayue2019; @liddell2018]. These types of analyses are routinely used in studies on MW [e.g., @aasen2024; @alexandersen2022; @drevland2024; @boayue2019; @boayue2020; @groot2022] and related fields [e.g., @filmer2019; @turi2019]. For each of the dependent variables MW, MB and S-MW, we included the predictors Time (scaled to 0-1), BV, AE, pre-task positive, pre-task negative and all of their respective interactions with the Sleep (continuous variable), and Sleep independently. For all participants, a random intercept was included in the model, and for the MB and S-MW models, we only considered those reports that were preceded by a MW report (score \>2) .

<!--# Models - Behaviour -->

For analysing task performance, we used two separate Bayesian hierarchical linear regression models for the dependent variables BV and AE. We included the predictors Time (time-on-task, scaled to 0-1), pre-task positive, pre-task negative, their respective interactions with Sleep (NS and PSD), and Sleep independently. We also included a random intercept for each participant. <!--# Models - Mood --> Lastly, we wanted to test whether sleep loss, test time (pre-task vs. post-task) and their interaction, would influence mood scores. To test these relationships, we ran two separate hierarchical Bayesian linear models for each mood score, positive and negative.

# Results

## Baseline Measures

Data from questionnaires collected at T0 are reported in @tbl-baseline-questionnaire. The participants' in our study did not indicate significant sleeping problems and are within the normal ranges of health individuals [in line with @saksvik-lehouillier2020], suggesting that the sample is from a health population. Three participants reported having night shift once per month (one participant did not provide an answer to this question). Most participants (`r sum(general_sleep_renamed$coff_day>0)`) reported to drink coffee, where `r sum(general_sleep_renamed$coff_day<4 & general_sleep_renamed$coff_day>0)` participants reported consuming 3 or fewer cups per day, while `r sum(general_sleep_renamed$coff_day>=4)` reported drinking 4 or more cups a day. As for the consumption of energy drinks, only `r sum(general_sleep_renamed$energy_drink>0)` participants reported consuming them, but maximum one per day.

```{r TBL - baseline questionnaire}
#| label: tbl-baseline-questionnaire
#| tbl-cap: Baseline Sleep, Mood and Beverage Consumption
#| tbl-cap-location: top
#| # Maybe add citations in apa-notes`? `
#| apa-note: ""
#| # META
#| include: TRUE
#| warning: FALSE
#| message: FALSE

outputs[["tbls"]][["baseline_sleep"]]
# citations
# [@krupp1989; jogns1991; @bastien2001; @buysse1989; @torsvall1980; @watson1988; @saksvik-lehouillier2020]
```

## Sleep

Participants self-reported and actigraphy-estimated sleep as well as sleep quality between the two conditions are presented in @tbl-sleep-times-sessions. We found overwhelming evidence that the sleep duration was reduced during the PSD period as compared to the NS period for all. It is noteworthy that, compared to the adjusted sleep, the total self-reported sleep length was longer while it was shorter relative to the actigraphy (see the Supplementary for a comparison). Interestingly, self-reported sleep quality was not affected by PSD (see the Supplementary). Furthermore, although participants were informed not to drink any caffeine-containing beverage before either of the experimental tests, `r demo_pre_test_sum |> filter(name=="no_caffeine" & condition=="control") |> pull(sum)` (`r demo_pre_test_sum |> filter(name=="no_caffeine" & condition=="control") |> pull(na)` missing) participants drank coffee before their NS tests, and `r demo_pre_test_sum |> filter(name=="no_caffeine" & condition=="SD") |> pull(sum)` (`r demo_pre_test_sum |> filter(name=="no_caffeine" & condition=="SD") |> pull(na)` missing) before their PSD tests. Nobody consumed alcohol before any of their test sessions, but `r demo_pre_test_sum |> filter(name=="no_alcohol" & condition == "SD") |> pull(na)` are missing from the NS condition and `r demo_pre_test_sum |> filter(name=="no_alcohol" & condition == "control") |> pull(na)` are missing from the sleep deprivation condition.

```{r TBL - sleep over session and modality}
#| label: tbl-sleep-times-sessions
#| tbl-cap: The Difference Between Sleep and Mood Across Sessions
#| tbl-cap-location: top
#| # META
#| include: true
#| message: false
#| warning: false

outputs[["tbls"]][["sleep_comparison"]]
```

## The Effects of PSD on the Thought Probes

Overall changes in probe responses between sleep conditions can be found in @fig-descriptives-probes.

```{r FIG - descriptive probes}
#| label: fig-descriptives-probes
#| fig-cap: "Changes in Probe Responses Between Conditions and Sample"
#| fig-cap-location: top
#| apa-note: "(a) Each subject is indicated by a triangle/circle and has a gray line connection them. Smaller transparent triangles/circles represent cases that are *excluded* form the reduced sample (see Supplementary). The black squares indicate the grand mean and standard error (SE) for the reduced sample (*n* = 23), while the black cross indicate the grand mean and SE for the whole sample (*n* = 34; i.e., including the smaller transparent cases). (b) Counts of each thought probe and their value between the samples. Note that the overall mean (a) and proportion (b) of each response barely change between sample, suggesing that the exclusion does not significantly impact the proportion of reported thought probe."
#| # META
#| include: true
#| message: false
#| warning: false
#| fig-format: svg
#| fig-height: 8
#| fig-width: 7

outputs[["figs"]][["probes__COMBINED_descriptives"]]
```

<!--# MW results -->

The Bayesian ordered-probit model for the dependent variable MW (@fig-probes-over-time left), indicated overwhelming evidence that MW increased with time after the NS (`r coef_hdi_text(c$mw[,"b_probenum_prop"])`), and increased with more variable responses (`r coef_hdi_text(c$mw[,"b_zlogbv"])`), but no evidence that randomness generation predicted MW (`r coef_hdi_text(c$mw[,"b_zlogapen"])`). Additionally, the model revealed overwhelming evidence that lower pre-task positive mood predicted less MW (`r coef_hdi_text(c$mw[,"b_pre_pos"])`), and weak evidence that pre-task negative mood increased MW (`r coef_hdi_text(c$mw[,"b_pre_neg"])`). After the PSD, the model indicated overwhelming evidence that participants generally reported more MW (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos"])`), but no substantial evidence supported an interaction with time-on-task (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"])`). Furthermore, very strong evidence indicated that sleep deprivation interacted with randomness generation (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:zlogapen"])`; sum: `r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:zlogapen"] + c$mw[,"b_zlogapen"])`), such that MW was predicted by less randomness with increasing amount of sleep deprivation. Strong evidence was found for an interaction between response variability and the PSD (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:zlogbv"])`; sum: (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:zlogbv"] + c$mw[,"b_zlogbv"])`), where increasing sleep deprivation weakened the relationship between BV and MW. No substantial evidence was found for an interaction with pre-task positive (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:pre_pos"])`), but some evidence was found for an interaction with pre-task negative mood (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:pre_neg"])`), where being in a more negative mood, after being sleep-deprived, attenuated its predictive effect on MW (sum effect: `r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:pre_neg"] + c$mw[,"b_pre_neg"])`.

The evidence for some of the coefficients were either attenuated or increased in the reduced model. After the NS, the pre-task positive mood indicated no substantial evidence for reduced MW (`r coef_hdi_text(r$mw[,"b_pre_pos"])`), while the pre-negative mood indicated overwhelming evidence that negative mood reduced MW (`r coef_hdi_text(r$mw[,"b_pre_neg"])`). After the PSD, no substantial evidence favoured a reduction in BV before a MW report (`r coef_hdi_text(r$mw[,"b_sleepdepSD:zlogbv"])`, and some evidence suggested that worse AE tended to precede a MW report (`r coef_hdi_text(r$mw[,"b_sleepdepSD:zlogapen"])`; but the sum effect indicated very strong evidence: `r coef_hdi_text(r$mw[,"b_zlogapen"] + r$mw[,"b_sleepdepSD:zlogapen"])`)

<!--# MB results -->

The corresponding model for the dependent variable MB (@fig-probes-over-time centre) indicated some evidence that MB increased with time-on-task after NS (`r coef_hdi_text(c$mb[,"b_probenum_prop"])`). MB was found to be positively predicted by response variability (`r coef_hdi_text(c$mb[,"b_zlogbv"])`), but no evidence that either randomness generation (`r coef_hdi_text(c$mb[,"b_zlogapen"])`), pre-positive (`r coef_hdi_text(c$mb[,"b_pre_pos"])`) or pre-task negative mood (`r coef_hdi_text(c$mb[,"b_pre_neg"])`) predicted MB. After the PSD, we found overwhelming evidence that participants generally reported more MB (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos"])`). On the other hand, no substantial evidence was found for either an interaction with time-on-task (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"])`), response variability (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos:zlogbv"])`), randomness score (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos:zlogapen"])`), pre-task positive (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos:pre_pos"])`), or pre-task negative mood (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos:pre_neg"])`).

In the reduced model, two coefficients had the same direction but indicated stronger evidence. Firstly, the model revealed stronger evidence that MB increased with time-on-task (`r coef_hdi_text(r$mb[,"b_probenum_prop"])`), and that more variable responses tended to occur before a MB report (`r coef_hdi_text(r$mb[,"b_zlogbv"])`). Otherwise, the effect of BV was attenuated to not substantial evidence (`r coef_hdi_text(r$mb[,"b_sleepdepSD:zlogbv"])` ).

<!--# S-MW results  -->

The model for the dependent variable S-MW (@fig-probes-over-time right), after the NS, indicated no evidence that S-MW increased with time-on-task (`r coef_hdi_text(c$smw[,"b_probenum_prop"])`), randomness score (`r coef_hdi_text(c$smw[,"b_zlogapen"])`), or pre-task negative mood (`r coef_hdi_text(c$smw[,"b_pre_neg"])`). Although, some evidence suggested that more variable responses tended to precede S-MW reports (`r coef_hdi_text(c$smw[,"b_zlogbv"])`), very strong evidence suggested that pre-task positive mood decreased S-MW reports in favor of D-MW (`r coef_hdi_text(c$smw[,"b_pre_pos"])`). After the PSD, no substantial evidence was found for a general change in S-MW (PSD: `r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos"])`), nor did it change with time-on-task (`r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"])`), randomness score (`r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:zlogapen"])`), or pre-task negative mood (`r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:pre_neg"])`). However, some evidence suggests an interaction with response variability (`r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:zlogbv"])`; sum effect: `r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:zlogbv"] + c$smw[,"b_zlogbv"])`), where (even) more variable responses tended to precede S-MW reports. Lastly, for the interaction with pre-task positive mood and PSD, we found overwhelming evidence pre-task positive mood no longer predicted S-MW tendency (`r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:pre_pos"])`; sum effect: `r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:pre_pos"] + c$smw[,"b_pre_pos"])`).

As for the reduced model, only weak evidence indicated that pre-task positive mood reduced the proportion of S-MW reports (`r coef_hdi_text(r$smw[,"b_pre_pos"])`), while stronger evidence was observed for an interaction between PSD and BV (`r coef_hdi_text(r$smw[,"b_sleepdepSD:zlogbv"])`), such that a S-MW tended to preceded by more variable responses.

```{r FIG - changes in probes over time}
#| label: fig-probes-over-time
#| fig-cap: Changes in Probe Responses Over Time
#| fig-cap-location: top
#| apa-note: "Interactions are related to the continuous model."
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-format: svg
#| fig-width: 6.5
#| fig-height: 4

outputs[["figs"]][["probes__PSD_x_time"]]
```

## The Effects of PSD on Task Performance

<!--#  Behavioural variability-->

The Bayesian hierarchical linear regression for the dependent variable BV (@fig-behaviour-over-time left), after the NS, indicated overwhelming evidence that response variability increased with time-on-task (`r coef_hdi_text(c$bv[,"b_probenum_prop"])`), but no substantial evidence that either pre-task positive (`r coef_hdi_text(c$bv[,"b_pre_pos"])`), or pre-task negative mood (`r coef_hdi_text(c$bv[,"b_pre_neg"])`) predicted changes in response variability. Following the PSD, the model revealed overwhelming evidence that participants had generally increased BV (`r coef_hdi_text(c$bv[,"b_c.Adjusted_Duration.diff.pos"])`), and strong evidence that variability further increased over time (PSD $\times$ Trial: `r coef_hdi_text(c$bv[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"])`). Furthermore, overwhelming evidence was found for an interaction with pre-task positive mood (`r coef_hdi_text(c$bv[,"b_c.Adjusted_Duration.diff.pos:pre_pos"])`), where higher positive mood scores, after the PSD, predicted reduced response variability even though the effect appeared to be rather small. Similarly for pre-task negative mood, a negative interaction was found, but only supported by some evidence (`r coef_hdi_text(c$bv[,"b_c.Adjusted_Duration.diff.pos:pre_neg"])`). In the reduced model, we only observed some evidence for the interaction between PSD and pre-task positive (`r coef_hdi_text(r$bv[,"b_sleepdepSD:pre_pos"])`).

<!--# Approximate entropy  -->

The corresponding analysis for the dependent variable AE (@fig-behaviour-over-time right), after the NS, indicated no evidence that randomness score changed with time (`r coef_hdi_text(c$ae[,"b_probenum_prop"])`) or pre-task mood scores (positive: `r coef_hdi_text(c$ae[,"b_pre_pos"])`; negative: `r coef_hdi_text(c$ae[,"b_pre_neg"])`). After the PSD period, no substantial evidence was found for either an initial change in randomness generation (`r coef_hdi_text(c$ae[,"b_c.Adjusted_Duration.diff.pos"])`), and no evidence that PSD interacted with pre-negative mood (`r coef_hdi_text(c$ae[,"b_c.Adjusted_Duration.diff.pos:pre_neg"])`). On the other hand, some evidence was found for an interaction with time as a function of increasing amounts of sleep deprivation (`r coef_hdi_text(c$ae[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"])`; sum: `r coef_hdi_text(c$ae[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"] + c$ae[,"b_probenum_prop"])`), where randomness was reduced over time with the magnitude of PSD. Furthermore, the model also revealed an interaction with pre-task positive mood (`r coef_hdi_text(c$ae[,"b_c.Adjusted_Duration.diff.pos:pre_pos"])`), where randomness score slightly increased with more pre-task positive mood. In the reduced model we observed strong evidence that PSD interacted with Trial (`r coef_hdi_text(r$ae[,"b_sleepdepSD:probenum_prop"])`), such that AE decreased over time more after participants were PSD.

```{r FIG - Changes in Behaviour over time by condition}
#| label: fig-behaviour-over-time
#| fig-cap: Changes in Behaviour Over Time
#| fig-cap-location: top
#| apa-note: ""
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-width: 5
#| fig-height: 3
#| fig-format: svg

outputs[["figs"]][["behaviour__PSD_x_time"]]
```

## The Effects of PSD on Mood

For the dependent variable positive mood, the analysis revealed overwhelming evidence that PSD reduced positive mood (`r coef_hdi_text(mood$p[,"b_c.Adjusted_Duration.diff.pos"])`), and overwhelming evidence that positive mood was reduced after the task (Pre-post: `r coef_hdi_text(mood$p[,"b_prepostpost"])`), but no evidence for an interaction between the PSD and Pre-post task (`r coef_hdi_text(mood$p[,"b_c.Adjusted_Duration.diff.pos:prepostpost"])`). As for the dependent variable negative mood, the analysis revealed strong evidence that PSD increased negative mood (`r coef_hdi_text(mood$n[,"b_c.Adjusted_Duration.diff.pos"])`), and no substantial evidence that the task influenced negative mood (`r coef_hdi_text(mood$n[,"b_prepostpost"])`), but strong evidence suggested that PSD and Pre-post task interacted (`r coef_hdi_text(mood$n[,"b_c.Adjusted_Duration.diff.pos:prepostpost"])`), such that after being the PSD they reported a decrease in negative mood after the task.

```{r FIG - Changes in Behaviour over time by condition}
#| label: fig-psd-on-mood
#| fig-cap: Changes in Mood Over Sleep Loss and Pre-Post Task
#| fig-cap-location: top
#| apa-note: ""
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-width: 5
#| fig-height: 3
#| fig-format: svg

outputs[["figs"]][["Mood__PSD"]]
```

<!--# The interactive effect of PSD on task performance on their associattion to MW -->

```{r FIG - MW interactions from model coefficients}
#| label: fig-interaction-behaviour-on-MW
#| fig-cap: The Interaction Between Behaviour and Sleep Loss on Their Relation to Mind Wandering
#| fig-cap-location: top
#| apa-note: "PSD = Partial sleep deprivation, NS = normal sleep. "
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-width: 5
#| fig-height: 3
#| fig-format: svg

outputs[["figs"]][["MW__PSD_x_behaviour"]]
```

```{r FIG - PSD x mood interacitons - BV + s-mw}
#| label: fig-psd-positive-interactions-bv+smw
#| fig-cap: The Interaction Between Behaviour and Sleep Loss on Their Relation to Mind Wandering
#| fig-cap-location: top
#| apa-note: "PSD = Partial sleep deprivation, NS = normal sleep. "
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-width: 5
#| fig-height: 3
#| fig-format: svg


outputs[["figs"]][["BV+SMW__PSD_x_pre_pos"]]

```

# Discussion

<!--# Summary -->

The current study examined how partial sleep deprivation (PSD) over three consecutive nights influenced MW, task performance and mood. The effect of PSD was pronounced, indicating a general increase in MW and MB, as well as higher BV, especially toward the end of the task. Randomness generation (AE) was not found to predict any of our thought probes after the normal sleep (NS) period, but after the PSD period, AE decreased with time-on-task and worse AE tended to precede MW reports. Furthermore, the PSD worsened mood by decreasing positive mood and slightly increasing negative mood. Additionally, we found some interactions between the mood variables and the thought probes, as well as between the mood scores and the behavioural indices. These results indicate the complex interplay between PSD, attention, executive functions and mood.

<!--# Sleep -->

Participants slept significantly less during the PSD period compared to the NS period, as confirmed by both the self-reported, actigraphy and adjusted sleep estimates. Although we observed great variation in the length of sleep duration between the participants, researchers have found that sensitivity to sleep loss varies between, but remains consistent within individuals [@hudson2020; @rupp2012; @vandongen2004; @dennis2017]. Therefore, participants might have adjusted their sleep duration according to their individual sensitivities to sleep loss. That is, some individuals might only need 1-hour less sleep to notice a detrimental effect, while others might need 3 hours less sleep to notice the same consequence. However, it could also be that the description of the PSD protocol was unclear. Specifically, the instructions stated that subjects were to go to sleep 3 hours later than *normal.* If participants had a variable bed time routine, what they should estimate their bed time as is not exactly clear. Although this did not appear to be a problem for @saksvik-lehouillier2020 and participants could always look at their own self-reported sleep. Clearer instructions are often implemented in many studies, such as a time-in-bed [TIB; e.g., @hudson2020; @vandongen2004] restriction. For example, a TIB of 5 hours indicates that, if participants wants to wake up at 08:00, they need to go to bed at 03:00. Thus, the variable sleep duration could result from individual differences in sensitivity to sleep loss or from the ambiguity in the PSD description.

<!--# We have some discrepancies, explain how these might arise -->

The self-report and actigraphy did not always provide similar sleep/wake times. The reason for the discrepancies likely stem from two sources: First, participants might have adjusted their reports to meet perceived study expectations (e.g., reporting later bedtimes or earlier wake times than actually experienced). Second, actigraphy has inherent limitations in accurately detecting sleep/wake transitions. For example, actigraphy might misinterpret motionless pre-sleep activities as sleep - such as using a mobile phone in bed, a common phenomenon in modern society [@gradisar2013; @hjetland2021; @hysing2015]. Similarly, it might fail to register actual wake times when participants remain in bed after waking. Regardless of these measurement challenges, our data showed that participants generally slept less during the PSD period compared to the NS period, as indicated by the self-report, actigraphy and most importantly, the adjusted sleep estimates that takes both measures into consideration.

<!--# Sleep quality -->

Sleep quality has been found to correlate positively with time awake, subjective sleep latency and objective sleep efficiency [@elmenhorst2008; @saksvik-lehouillier2020]. Although our participants reported sleeping less, they did not report better sleep quality. Given the variation in sleep reduction between participants, sleep quality might have improved only in participants who slept even less. However, an exploratory analysis (see the Supplementary) did not find support for this idea: Those who slept the least had similar sleep quality as those who slept more, indicating that the PSD did not influence sleep quality in our sample. One alternative explanation is that participants used their phone before falling asleep, which could have reduced their sleep quality [e.g., @exelmans2019; @hjetland2021; @hysing2015; @lastella2020; @nosetti2021], possibly through blue-light [@chang2015]. <!--# Thus, it could be that  during the PSD but was reduced but was cancelled by  and reduced from phone use, they would ultimately cancel each other out. sleep quality was improved during the PSD, but that improvement could be cancelled by their phone use.  --> However, this alternative explanation is speculative as we do not have data on participant's phone use, and future research is needed to confirm whether bedtime phone use impedes improvements in sleep quality following sleep deprivation.

<!--# PSD on Thought probes -->

The amount of PSD increased reports of MW and MB, but not S-MW. In line with previous research, participants reported more MW [@poh2016; @marcusson-clavertz2019; @cárdenas-egúsquiza2022b] and MB [@jubera-garcía2021sleepMW; @boulakis2024]. MW was increased with time [@zanesco2024b; @aasen2024; @alexandersen2024; @boayue2020] that was further exacerbated by increasing amounts of PSD. MB, however, was somewhat increased with time, but greatly attenuated with increasing amounts of PSD. This attenuation might have arisen due to the brain's adaptive ability to down regulate cognitive processing capacity to maintain stable cognitive processing over time [@belenky2003]. Contrary to our hypothesis, we did not find S-MW to be increased with increasing amounts of PSD. Researchers have demonstrated that motivation [@seli2015], EF [@mcvay2009] and task difficulty [@seli2016; @seli2018] can influence S-MW tendency. PSD might have reduced motivation, increasing D-MW [@seli2015], while it could also have reduced EF [@saksvik-lehouillier2020; @skurvydas2020], increasing S-MW [@mcvay2009]. These effects would be counteractive, ultimately leaving S-MW tendency unchanged.[^2] In summary, MW and MB, but not S-MW was increased with increasing amount of PSD.

[^2]: It is also possible that the task could have been perceived as more difficult due to the PSD's effect on EF, which could have caused more S-MW [@seli2016; but see @seli2018].

<!--# PSD on Behaviour -->

Sleep deprivation showed the expected effect for BV but not AE. BV was increased with increasing amounts of PSD, that also further exacerbated with time. These results align with previous research demonstrating that sleep loss disrupts the stability of cognitive performance [@jackson2013; @doran2001], resulting in greater fluctuations in response timing that further increases with time-on-task [@thomson2014; @zanesco2024a; @zanesco2024b; @alexandersen2024]. However, contrary to our hypothesis, PSD did not impact overall AE performance, but some evidence suggested that AE decreased with time. The fact that AE was not influenced by the PSD is surprising given that executive functions are typically considered vulnerable to sleep deprivation [@alhola2007] and previous research has found reduced randomness generation following TSD [@heuer2005]. One study has suggested that AE requires sustained periods of stable performance to derive its predictive power [@alexandersen2024] and it is possible that such periods did not consistently occur in our sample. However, the reduction in AE over time suggests that EF were influenced by the PSD, possibly by reducing executive resources [e.g., @thomson2015]. The reduction in executive resources would have made EF more vulnerable which could reduce its overall performance. Regardless of the precise relationship between PSD and AE, we found the expected increase in BV following PSD.

<!--# PSD ->  Mood -->

With respect to the effect of PSD on mood, we observed the expected patterns. Participants reported significantly reduced positive and slightly increased negative mood [@lo2016; @saksvik-lehouillier2020; @palmer2024]. Despite the PSD-associated increase of negative mood before the task, this effect was relatively small, and disappeared after the task. For positive mood, we found the same decrease from pre- to post-task in both conditions, suggesting that the PSD did not differentially impact the effect of engaging in the task. These findings align with the emerging view that positive and negative affect may be differentially sensitive to sleep loss, with positive emotions showing greater vulnerability [@palmer2024]. This asymmetrical effect on emotional experience could have important implications for daily functioning, as reduced positive affect might diminish motivation [@seli2015; @erez2002], creativity [@baas2008], and social engagement [@watson1988] even when negative emotions are not markedly increased. Moreover, the discrepancy might be of greater concern for the development of various psychological disorders (e.g., major depressive disorder) because the loss of positive mood could predispose individuals to think in a less flexible [@marchetti2016] and more rigid manner [@ottaviani2013; @ottaviani2015].

<!--# Mood x PSD -> behaviour + attention  -->

Mood also had interesting relations to behaviour and the thought probes. We observed a negative relationship between pre-task positive mood and MW \[@smallwood2009\] independent of the PSD, but no relationship with negative mood \[@fell2024\]. In addition, we observed a negative relationship between pre-task positive mood and S-MW [@seli2015], but only after the NS period. This pattern indicates that participants in a more positive affective state might have subjectively experienced more control, even though they did not indicate improvements in their objective performance. After the PSD, however, we observed that pre-task positive mood decreased response variability (improved *performance*), but also removed the increase in D-MW relative to S-MW. Previous research has demonstrated that increased positive mood can increase working memory capacity [@ribeiro2019; @yang2013; but see also @mitchell2007], however, we did not observe any effects of mood on sequence randomness. These patterns suggests that positive mood might be related to motivation [@erez2002] which improve attentional focus to the task, and only under conditions of reduced EF does motivation improve response variability.

<!--# Behaviour x PSD -> Thought probes -->

Regarding the associations between behaviour and the thought probes, we found some of the expected effects. After the NS, we observed that BV tended to be increased before MW episodes as observed elsewhere [e.g., @aasen2024; @alexandersen2025; @boayue2020; @seli2013b], and of those episodes, those that were blank tended to have higher BV than those with content, in line with previous research[@andrillon2021; @boulakis2024]. Intriguingly, after the PSD, we observed evidence for a slight reduction between the BV-MW relationship, while the BV-MB relationship was increased, and a positive relationship between BV and S-MW was observed. The reduced relationship between MW and BV might be observed because BV is not a perfect predictor of MW, and other factors might be impacted by the PSD, such as motivation [@engle-friedman2003; @killgore2010] and specific brain regions \[e.g., supramarginal gyrus, @groot2022; @drevland2024\]. The stronger relationship between BV and MB could be observed due to the increased MB reports after the PSD, which heighten the power to detect an effect. As for the relationship between BV and S-MW, it might be that D-MW only occurs when sufficient executive resources are available [@robison2018], or if the task is perceived as easy [@seli2016], or when motivation is low [@seli2015; @robison2018], or a combination of them. In summary, we observed some changes in the relationship to BV and the thought probes as a function of the PSD.

As for the relationship between the thought probes and randomness generation, we only found some expected relationships after the PSD. Following the NS period, none of the thought probes were predicted by AE, in contrast previous research [@aasen2024; @boayue2020; @groot2022; @drevland2024]. However, after the PSD, AE tended to reduce with time, and we observed a negative relationship with MW. These patterns suggests that the PSD may have created the conditions where declining executive control [or executive resources, e.g., @thomson2015] becomes more tightly coupled with attentional lapses. The interaction with AE can be interpreted in light of the resource-control account [@thomson2015], where the PSD might reduce the availability of cognitive resources [@alhola2007]. Then, when shifts of attention occur, they draw away the sparsely available resources from the task that would reduce EF. Indeed, research suggests that sleep deprivation reduces the activity in the fronto-parietal control network [FPCN; @yao2023], an important network for EF [e.g., @dosenbach2008], that has also been implicated in MW [@christoff2009; @groot2021; @dixon2018]. In sum, AE was negatively related to MW only after the PSD, plausibly through reduced resources availability following the PSD.

<!--# Difference between the models -->

{{{This might fit better in the Supplementary?}}}

It is important to note that not all effects described here were equally convincing in the reduced sample (see the Supplementary). However, and importantly, the effects described here were *all* pointing in the same direction as in the reduced model, and were mostly related to changes in mood effects – which are particularly sensitive to the sample size. To highlight one change between the models, in the MB model regarding the BV effect, we found weak evidence in the reduced sample model, but strong evidence in the full sample model, still in the positive direction. <!--# To illustrate, in the MB ordered-probit model, only weak evidence suggested that BV tended to precede a MB reports, while only some evidence suggested that positive mood reduced MW, S-MW, and BV. Interestingly, in the reduced MW probit model, we found very strong evidence that negative mood predicted more MW, in line with other works [@killingsworth2010; @smallwood2009], and, in the AE linear model, we observed strong evidence PSD reduced AE over time. --> It is also important to point out that we here treated the PSD as a continuous variable, while in the reduced sample it was treated as dicotomous variable with a more strict threshold. The observed changes could result from the more solid effect of the PSD, but could also related to the reduced sample. Importantly, the differences are not drastic and only related to changes in the *degree of evidence* - no predictors reversed its effect. Nevertheless, caution should be exercised when interpreting the mood effects, and future research will be necessary to confirm these findings.

<!--# Future research -->

While the PSD protocol employed here is more ecologically valid than TSD, it is nevertheless a rather short-term intervention. Previous research has investigated longer-term effects of sleep restrictions, such as a week [@dinges1997] and 14 days [@vandongen2003]. However, chronic sleep deprivation can be stretched over multiple weeks and months, even with a very short reduction of sleep [e.g., 30 minutes, @sivertsen2019]. Such long-term experiments would be more ecologically valid, but more difficult, unpractical and inconvenient. Although, some research suggests that there are overlaps between the effects of PSD and TSD [@vandongen2003], much research suggests that these two conditions are dissociable [@belenky2003; @banks2007; @lo2016]. For instance, @belenky2003 suggests that PSD causes adaptive changes in the brain to maintain stable performance by down-regulating processing capacity. However, this adaptation also comes at the cost of requiring longer recovery, probably because the down-regulatory adaptations need to be unwound. Future research may want to investigate exactly how many nights of reduced sleep is necessary to induce these adaptive changes, and how many nights are needed for full recovery.

Furthermore, some research indicates that a variable sleep schedule might have a greater influence on performance than the length of sleep [@sletten2023; @ness2018]. Thus, if our sample maintained a relatively stable sleeping routine, the detrimental effects of the PSD might have been reduced compared to having a more variable sleep schedule. The fact that an inconsistent sleep schedule might impact performance more than sleep duration could be related to shifts in the circadian rhythm [@phillips2017] which can disrupt optimal recovery. Future research may want to investigate how variation in sleep timing impacts performance, especially as sleep variation is a common occurrence in certain populations [e.g., @phillips2017].

## Summary

We investigated how 3 nights of partial sleep deprivation influenced attentional states, behavioural performance, and mood. Participants slept on average 1.7 hours less for three nights and led to increased mind wandering, mind blanking, response variability, increased negative mood, decreased negative mood and worse randomness generation over time. We observed that worse sequence randomness tended to precede MW reports only after the PSD, suggesting that the PSD might have reduced executive resources [@heuer2005]. This reduction, would have caused a MW episode to consume the few remaining executive resources [@thomson2015], leaving fewer resources available for the generation of randomness. Our results add to the growing body of literature demonstrating that even modest reductions in sleep [@saksvik-lehouillier2020] can significantly impair sustained attention, which are especially concerning for safety-critical activities such as driving [@rodd2017; @yanko2014].

# References
