---
title: "The Effects of Partial Sleep Deprivation on Mind Wandering, Behaviour and Mood"
shorttitle: "PSD on MW, EFs and Mood"
authors:
  - name: Steffen Rygg Aasen
    orcid: 0000-0003-2807-7504
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - writing
      - data curation
      - investigation
      - formal analysis
      - resources
      - visualization
  - name: Samy Engebø Babiker
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - data curation
      - investigation
      - resources
      - editing
  - name: Ingvild Saksvik-Lehouillier
    orcid: 0000-0002-1953-7364
    affiliation: Norwegian University of Science and Technology
    roles: 
      - resources
      - methodology
      # - editing
  # - name: Julia W. Y. Kam
  #   orcid: 0000-0002-2369-2148
  #   affiliation: Helen Wills Neuroscience Institute, University of California – Berkeley, Berkeley, CA, USA
  #   roles: 
  #     - editing
  - name: Aurora Marie Kleivik Vangen
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - data curation
      - investigation
      - resources
  - name: Silje Balsnes Haugan
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - data curation
      - investigation
      - resources
  - name: Sanna Aurora Sandell
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - data curation
      - investigation
      - resources
  - name: Gábor Csifcsák
    orcid: 0000-0001-7719-7073
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - conceptualization
      - methodology
      - supervision
      - validation
      - resources
      - editing
  - name: Matthias Mittner
    orcid: 0000-0003-0205-7353
    affiliation: Department of Psychology, UiT - The Arctic University of Norway
    roles: 
      - conceptualization
      - funding acquisition
      - project administration
      - methodology
      - software
      - resources
      - supervision
      - validation
      - visualization
      - editing
    corresponding: true
abstract: "This study investigates the effect of partial sleep deprivation (PSD) over 3 consecutive nights on our tendency to shift our attention, behaviour and mood. We used a within-subject cross-over design, where 34 healthy adults engaged in the finger-tapping random sequence generaiton task (FT-RSGT) after a period of PSD and normal sleep (NS). The participants indicated to sleep on average 1.7 hours less during the PSD period compared to the NS period. The FT-RSGT provides two objective indices of attention, behavioural variability (BV) and randomness generation (AE), and subjective reports of attention. We found that PSD increased mind wandering, mind blanking, BV, decreased positive mood and slightly increased negative mood, meanwhile AE was generally unaffected by the PSD. These findings contribute to our understanding of how even moderate sleep reduction affects attention, highlighting the possibly severe consequences of moderate sleep loss in safety-critical activities requiring sustained attention, such as driving. Additionally, these results are problematic granted the rising social trend towards less sleep duration, and suggest that only a couple of days might be sufficient to significantly impair our everyday behaviour."
keywords: [attention, sleep, mind wandering, partial sleep deprivation, executive functions, mood]
author-note:
  disclosures:
    conflict-of-interest: "The author has no conflict of interest to declare."
    data-sharing: "Data from this study can be accessed at the Open Science Foundation (OSF): https://osf.io/xq6wr/."
    financial-support: "WHAT ARE WE SUPPORTED BY? "
    gratitude: "We would like to extend our gratitude to Ragnhild Nicolaisen Drevland for helping in the data collection. Moreover, we would like to express our deep gratitude to all the participants slept less (for Science!)."
bibliography: references.bib

# Format
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf: default
  apaquarto-typst: default

# Code settings
editor: source
execute:
  include: false
  echo: false
---

```{r load project data munge Parameters, warning=FALSE}
relative_path <- "../" # b/c quarto 
source( paste0(relative_path, "lib/quarto_load_project.R") )
quarto_load_project(relative_path, load_data_data = T)
```

# Introduction

Sufficient sleep is paramount for optimal functioning in our everyday lives [@alhola2007] and insufficient sleep demonstrates far-reaching consequences across many different health outcomes, such as depression and obesity [@chattu2019]. Moreover, numerous studies demonstrate the negative influence of sleep restriction on sustained attention, executive functions, emotional stability, and memory consolidation [@åkerstedt2007; @alhola2007; @anderson2011; @lo2016; @poh2016; @stickgold2007]. For example, sleep deprivation increases attentional lapses and impulsivity during sustained attention tasks [@anderson2011; @lo2016; @saksvik-lehouillier2020], and decreases speed of processing during an arithmetic task [@lo2016]. Insufficient sleep also impedes memory consolidation, likely, by disrupting neural plasticity processes that occur during sleep, especially those associated with slow-wave sleep and rapid-eye-movement sleep [@stickgold2007] {{{I suppose talking about memory consolidation is a bit beside out research here, and probably best left aside to not distract the reader? }}}. Additionally, sleep loss can reduce positive [@lo2016; @saksvik-lehouillier2020] and increase negative emotions [@cárdenas-egúsquiza2022b] influencing our overall well-being in a way that might predispose individuals towards depression [@chaieb2022].

Most studies implement total sleep deprivation (TSD) where participants who are not allowed to sleep at all are contrasted with those following their regular sleep schedule. However, TSD is less common compared to partial sleep deprivation [PSD; e.g., @sheehan2019; @sivertsen2019], where individuals sleep less than the recommended duration (e.g., an adult sleeping 5 hours instead of the recommended 7-8 hours), an effect that can accumulate over multiple days. Intriguingly, the accumulating effect of PSD might impact cognitive functions and mood more than TSD [@pilcher1996; @salfi2020]. In general, PSD may be less noticeable than TSD because people can habituate to the reduced amount of sleep [@balkin2008] and develop coping strategies to compensate for the reduced cognitive functioning [@rupp2009]. Moreover, PSD appears to require multiple days of normal sleep (NS) to recover normal cognitive functioning whereas a single night is often sufficient after TSD \[@balkin2008; @belenky2003; @lo2016; @cohen2010\]. In summary, despite the prevalence of PSD [e.g., @sheehan2019; @sivertsen2019], its effects remain surprisingly underexplored [@alhola2007].

Insufficient sleep influences most cognitive functions, including attention. Sleep deprivation worsen task focus by increasing shifts of attention away from the task to internal thoughts, a phenomenon researchers usually call mind wandering [MW; @smallwood2015]. For example, one study indicated that one night of TSD increased MW [@poh2016]. Additionally, recent evidence suggests that the effect between MW and sleep might be directional. Two diary studies found that self-reported sleep disturbances predicted increased MW the following day [@cárdenas-egúsquiza2022b], while more MW during the day did not predict sleep problems the following night [@marcusson-clavertz2019; but see @fell2024].

Given the causal evidence showing that sleep deprivation increases MW, MW might be one of the underlying consequences of reduced cognitive performance following sleep deprivation. Indeed, similar behavioural consequences are often observed following sleep deprivation and MW episodes. For example, in a simulated driving environment, participants show greater driving speed variation, lane variation, and braking response times following sleep deprivation [@jackson2013]. Similarly, MW during a simulated driving environment increase driving speed and braking response times [@rodd2017; @yanko2014]. Furthermore, during more formal cognitive tests, sleep deprivation has been found to increase response variability [@jackson2013] and reduce random sequence generation [@heuer2005], similar to how MW episodes are related to increased response variability [@boayue2020; @aasen2024; @alexandersen2022; @baldwin2017; @seli2013b] and reduced random sequence generation [e.g., @drevland2024; @teasdale1995].

The tendency to shift our attention is multifaceted and complex. For example, MW episodes typically have a specific reportable content, like thinking about what to make for dinner or upcoming social events. Occasionally, however, we might find ourselves not focusing on the task but at the same time not thinking about anything. This lack of content during inattention is referred to as mind blanking (MB), and proposals have suggested that MB as well as MW might be increased by sleep deprivation [@andrillon2019; @jubera-garcía2021sleepMW] - something that was recently confirmed [@boulakis2024]. Furthermore, research has demonstrated that MB is associated with slower response times, increased misses [@andrillon2021; @mortaheb2022], reduced pupil size, and increased subjective sleepiness [@unsworth2018] compared to MW. These associations suggest that MB may be linked to reduced information processing [@boulakis2023; @musat2024; @mortaheb2022] plausibly through deeper task disengagement [i.e., stronger perceptual coupling; @smallwood2015; e.g., @ward2013] and lower arousal [@boulakis2024]. Altogether, these findings indicate that MB can be distinguished from MW, and that both are influenced by sleep deprivation.

The tendency to shift our attention can be done deliberately or happen spontaneously. Research suggests that participants with lower motivation report more MW and are engaged in more deliberate (D-MW) than spontaneous MW [S-MW; @robison2018; @seli2015]. Moreover, participants seem to engage more in S-MW than D-MW during difficult tasks, and, conversely, more D-MW than S-MW during easy tasks [@seli2016]. Additionally, research has demonstrated a negative relationship between working memory capacity (WMC) and S-MW [@ju2018; @robison2018] suggesting that WMC plays a role in regulating MW intention. Given the negative impact of sleep deprivation on both executive functioning [EF; @heuer2005] and motivation [@engle-friedman2003; @killgore2010], it is plausible that sleep-deprived individuals may experience alterations in both their capacity and intention to engage in different forms of MW.

Variation in sleep has also been related to variations in mood. For example, sleep-deprived individuals tend to reported less positive and more negative mood, while also indicating more anxiety and depressive symptoms [@palmer2024; @short2020]. Similarly, MW has also been associated with negative affect in the general population [@killingsworth2010], and is increased in patients suffering from various psychiatric disorders, such as depression and anxiety disorder [@marchetti2012; @marchetti2016; @smallwood2007b]. Moreover, it has been proposed that positive mood might be related to motivation [@erez2002], which has been shown to be reduced after sleep deprivation [@engle-friedman2003] and can influence MW tendency [@seli2015]. Thus, it is possible that the changes in mood following sleep deprivation could influence the tendency to mind wander and its various manifestations.

As noted, EF is negatively influenced by sleep deprivation [@heuer2005], yet much research on MW does not adequately load executive functions [@boayue2020]. The finger tapping random sequence generation task [FT-RSGT; @boayue2020], however, effectively measures EF while also capturing MW tendencies. The FT-RSGT requires participants to create a random sequence of responses using two dedicated keys, while matching the rhythm of an ongoing metronome as closely as possible. Producing random sequences engages the central executive [@baddeley1998], by requiring inhibitory control of pre-potent responses, as well as mental monitoring and updating of previously generated sequences, all of which are core functions of the executive system [@miyake2000]. The randomness of the sequences has been negatively related to MW [e.g., @teasdale1995; @boayue2020; @aasen2024; @groot2022], while response variability has been positively associated with MW [@boayue2020; @alexandersen2024; @bastian2013; @cheyne2009; @seli2013b].

To unravel the complex relationship between MW (and its various manifestations), the executive system, mood, and sleep, we conducted a within-subject, cross-over study using a PSD protocol [@saksvik-lehouillier2020]. Our main hypotheses concerned the effect of PSD on MW and FT-RSGT performance. Specifically, we expected that PSD would increase participants’ overall propensity to engage in MW, increase the frequency of MB episodes relative to MW with reportable content, lead to more S-MW relative to D-MW, increase negative mood, reduce positive mood, increase BV and reduce randomness generation in the FT-RSGT. Additionally, we expected that PSD-associated reduction in mood before the task would predict the amount of MW during the task. Moreover, we also aimed to assess how PSD would influence the interaction between measures of EF and different forms of MW, and whether PSD and changed mood scores would moderate the relationship between task-focus and EF performance.

# Method

## Participants

We recruited `r subj$full$num` healthy adults with an average age of `r subj$full$age` (*SD* = `r subj$full$age_sd`, range = `r subj$full$age_min`-`r subj$full$age_max`, females = `r subj$full$gender_f`) to a within-subject cross-over design study {{{This is noted in the intro}}}. In the analyses reported here, no participants were excluded. In the Supplementary, however, we report analyses for a reduced sample (see below) where `r subj$red$exc` participants were excluded, leaving us with `r subj$red$num` participants with an average age of `r subj$red$age` (*SD* = `r subj$red$age_sd`, range = `r subj$red$age_min`-`r subj$red$age_max`, females = `r subj$red$gender_f`). Data-collection took place between March 2021 and May 2022.

The experiment was approved by the institutional ethics board at UiT The Arctic University of Norway. Participants were recruited via social media posts, flyers, and personal contacts in Tromsø. To be included, participants needed to be between 18 and 50 years, with no history of or ongoing mental illness, as well as not taking any psychoactive drugs. Participants were also asked whether they had any hearing impairments (used to adjust the metronome sound, but none reported any impairments). Lastly, participants had to give written consent to participate, and they were compensated for their time with a gift card at a local shopping mall valued at 500 NOK (approximately 50 USD).

## The Task

The task (@fig-protocol-a) was coded in PsychoPy, Standalone version 1.83.04_win32 [@peirce2007]. The metronome had an inter-stimulus interval (ISI) of 750 ms where it played a sound of 440 Hz for 75 ms on two stereo speakers placed beside the monitor, approximately 0.5 meter from the participant, with a peak volume of approximately 60 dB. Furthermore, participants were pseudo-randomly probed about their attentional state. The probes were presented randomly every 40 to 80 seconds. Each time, participants were asked three consecutive questions relating to their attention, which were answered on a 4-point Likert scale. The first question was related to their focus: “To what degree were you focused on the task just before this question?” (1: “clearly unfocused”, 4: “clearly focused”). The second question was related to the content of the MW episode: “To the degree to which you were not focusing on the task, were you thinking of something in particular or just thinking about nothing?” (1: “clearly nothing”, 4: “clearly something specific”). Lastly, we wanted to distinguish between the intentionality of the MW episode: “Were you deliberate about where you focused your attention (either on-task or elsewhere) or did it happen spontaneously?” (1: “clearly spontaneous” to “clearly deliberate”). The task lasted for approximately 25 minutes answering a total of 25 probes. All material in this study is publicly available in Norwegian at the Open Science Framework (OSF; <https://osf.io/xq6wr/>; partial English translations can be found at <https://osf.io/cv24f/> and <https://osf.io/txpu2/>).

Both BV and AE were estimated from the 25 stimuli preceding a thought probe, corresponding to 18.75 seconds [@boayue2020]. BV was calculated as the log-transformed standard deviation of the inter-tap-intervals and then z-scored across all subjects. The AE variable [@pincus1997] was parametrized by parameter *m* that indicates the maximum length of subsequence’s used for the entropy-calculation. For the current analysis, we used *m* = 2, which has been found to provide the most stable estimates with a robust effect in relation to MW [@boayue2020]. Raw AE values were transformed using $-log(log(2)-AE)$ and then z-scored across all subjects. Higher values of AE imply greater irregularities (more random responding) of the sequence [@pincus1991], associated with stronger recruitment of executive resources [@teasdale1995].

<!-- ::: {#fig-protocol layout-nrow=2 fig-cap-location="top"} -->

![Finger-tapping random sequence generation task (FT-RSGT)](../outputs/experiment/FT-RSGT_overview.svg){#fig-protocol-a width="60%"} <!-- !["Sleep protocol"](../outputs/experiment/sleep_dot.svg){#fig-protocol-b width=50%} --> <!-- ::: -->

## Sleep Paradigm

<!--# Paradigm in general  -->

To investigate the effect of PSD on MW and executive functions, we used a modified PSD protocol described by [@saksvik-lehouillier2020; @fig-protocol-b]. In their study, the authors asked their participants to sleep 2 hours less than usual each night for 3 consecutive nights, and their participants achieved an average of 2.23 (*SD* = 0.72) hours less sleep according to their actigraphy measure. To increase the effectiveness of the PSD manipulation [@lim2008; @saksvik-lehouillier2020; @schwarz2016], we asked our participants to sleep 3 hours less for three consecutive nights in the PSD condition. During normal sleep, participants were instructed to not go to sleep later than 00:30 and not wake up later than 09:00. Meanwhile, during the PSD, they were instructed to go to sleep 3 hours later than usual, but still wake up at the same time as during their normal sleep

<!--# Setup of the protocol -->

The order of the conditions was counter-balanced across subjects, resulting in two groups: Participants in the early sleep deprivation group (ESD) underwent PSD before the first data-collection, while participants in the late sleep deprivation group (LSD) had a period of controlled NS before the first data collection (see @protocol-fig). Participants were pseudo-randomly assigned to either the LSD or ESD condition, but we were flexible with group assignment to better fit the participants' personal schedule. Data from both groups were merged for the analysis.

!["Sleep protocol"](../outputs/experiment/sleep_dot.svg){#fig-protocol-b width="50%"}

<!--# Measuring the compliance -->

To ensure compliance with the sleep protocol, we used two measures of participants' sleep. Firstly, participants were required to wear an Actiwatch Spectrum Pro (Philips Respironics, USA; <https://tinyurl.com/yjr2szxg>) during their sleep, but were recommended to wear it all the time. The Actiwatch measures physical activity and ambient light which is used to estimate participants' sleep- and waking-time, sleep latency and total sleep using the accompanying software, Philips Actiware 6.0.2. Secondly, participants filled out a sleep diary in the morning after waking. They were asked when they went to bed, when they tried to fall asleep, how long it took before they fell asleep, how many times they woke up during the night before waking up in the morning, how long these wake-ups were, when the last wake-up occurred, when they rose from the bed, to rank their sleep quality from very bad to very good (5-point Likert scale), and if they took off the Actiwatch at any time the day before. To calculate the participants’ sleep duration, we first estimated their sleep onset by adding their self-reported bedtime and their estimated sleep delay. Participants' wake times were defined as their self-reported last waking time, and then, their sleep duration was calculated by taking the difference between their sleep onset and their wake time. For all sleep estimates, we used the 3 nights preceding each test because the PSD condition was limited to 3 nights of sleep restriction.

<!--# Discrepancies between the measures -->

The self-reported and actigraphy estimated sleep onset and wake times did not always converge (see @fig-SD-between-modalities). Cases were considered diverging when they differed by more than 15 minutes. To resolve these cases, we loosely followed the procedure described by @follesø2021 to estimate an adjusted sleep duration. The procedure provides a standardized, hierarchical method to redefine rest intervals before falling asleep and after waking up in the morning, by systematically comparing algorithm-generated estimates with event markers, sleep diary entries, and changes in activity and light patterns.

<!--# The Shiny application -->

To re-evaluate the divergent cases, we created a Shiny application. The application visualized time on the x-axis and the actigraphy recorded activity (raw) and light (log transformed) on separately y-axis. Coloured marks highlighted the self-reported and actigraphy sleep onset and wake time. The rest interval was indicated through shaded areas ranging from the sleep attempt to the sleep onset, and the wake time to the rise time for both modalities. Two sliders were provided (hours and minutes) to indicate the new onset/wake time, but separate buttons were provided to apply the exact values of the self-reported or actigraphy onset/wake times.

<!--# Resolving the discrepancies -->

We had two researchers (S.R.A. and G.C.) individually reevaluate the sleep in general terms with the procedure by @follesø2021. After the evaluation, there was a high level of agreement between the researchers (wake *r* = `r sleep_adjustment_round1_df |> filter(cond=="wake") |> pull(cor) |> fmt_APA_numbers(.low_val = T)`, onset *r* = `r sleep_adjustment_round1_df |> filter(cond=="onset") |> pull(cor) |> fmt_APA_numbers(.low_val = T)`). However, to reach full agreement, the researchers collectively evaluated those cases diverging by more than 15 minutes of those cases they had evaluated, while cases below 15 minutes were averaged. Importantly, since we incorporated both the actigraphy and self-report, we relied on either type (e.g., self-report) if a report from the other (e.g., actigraphy) was missing. Thus, the adjusted sleep benefited from a more complete dataset for the sleep estimation. For our analyses, we chose to use the adjusted sleep estimates, as it resolved some obvious discrepant cases. Importantly, we treated the PSD variable as a *continuous*, rather than as a dichotomous variable (condition), relating to the relative sleep difference between the NS and the PSD condition for each subject respectively.

```{r FIG - sleep deprivation between source type}
#| label: fig-SD-between-modalities
#| fig-cap: Difference in Sleep Duration Between Sources 
#| fig-cap-location: top
#| apa-note: "Sleep deprivation compared to the normal sleep (NS). Sleep duration is averaged for the three nights preceding each test, and the difference is caluclated as partial sleep deprivation (PSD) - NS. Black solid line indicate the original acceptance threshold, while the red dashed line indicated the adjusted sleep threshold (the criterion used for the reduced sample). Cases above the 1.5 threshold are highlighted by their associated participant code."
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-format: svg
#| fig-width: 10

outputs[["figs"]][["PSD__report_type"]]
```

# Procedure

### Baseline

At the start of the study (T0) participants were informed about the experimental protocol and had to give written consent before proceeding. Once written consent was given, participants were handed out a Phillips Actiwatch, a sleep diary, and a general questionnaire about their sleep patterns. In the general sleep questionnaire, we asked whether the participants had any night shifts, if they worked, with four discrete options: (1) nothing, (2) one night or less a month, (3) 2-4 nights per month or (4) 2 nights per week or more. Furthermore, we asked about the participants’ sleep preference during the weekdays and weekends: when they preferred to fall asleep, wake up, and for how long they preferred to sleep. They were then asked about how many hours they slept in a typical 24-hour period during the weekdays and weekends. Lastly, we inquired how many caffeine-containing beverages (e.g., coffee, tea) and energy drinks they usually consume in a normal day [@saksvik-lehouillier2020].

{{{MM: The below can be moved to the supplementary, and a shorter summary can be written, if space becomes an issue. }}}

We used verified scales to measure sleep, fatigue, insomnia, and sleep quality [@saksvik-lehouillier2020]. Participants' fatigue was measured with the Fatigue Severity Scale [@krupp1989] with lower scores indicating less fatigue. To measure sleepiness, we used the Epworth Sleepiness Scale [@johns1991], where lower scores (0-5) indicate low daytime sleepiness while higher scores (16-24) indicate severe excessive daytime sleepiness. Sleeping problems were measured with the Insomnia Severity Index [@bastien2001] where lower scores (0-7) indicate no significant insomnia, while higher scores (22-28) indicate severe insomnia. Participant’s sleep quality was measured with the Pittsburgh Sleep Quality Index [@buysse1989] that inquires about participants’ sleep time, sleep delay, wake time and sleep amount, where higher scores indicating worse sleep quality. Participant’s diurnal preferences were assessed with the Diurnal Scale [@torsvall1980] where higher scores indicate a preference to wake up and go to sleep early (‘early birds’), and lower scores indicate a preference to go to sleep later and wake up late (‘night owls’). To measure the participant’s mood, we used the Positive and Negative Affect Schedule [PANAS, @watson1988] where higher scores (min 10, max 50) indicate more intensive emotional states (for our analyses the PANAS was rescaled to range from 0 to 40). Lastly, we also asked participants about their alcohol consumption habit, derived from [@saksvik-lehouillier2020]. This consisted of 3 different questions with 5 answer alternatives (ranging from low to high consumption).

### Data Collection

Data collection procedures[^1] were identical for the LSD and ESD groups. LSD participants came to the first session after 7 days of NS, while in the ESD group, the session commenced after 3 additional days of PSD.

[^1]: Due to the SARS-CoV-2 pandemic in 2020-2022, we followed the safety guidelines given by our faculty. All surfaces touched by the participants were cleaned after each session, the experimenters wore disposable protective gear (e.g., gloves and facemasks), and participants were provided with a facemask if they did not already wear one. During testing, participants could take off their facemasks, but had to put it back on after the test was completed.

When participants arrived in the laboratory, they completed the PANAS to assess pre-session mood and were provided with the instructions for the FT-RSGT. During the experiment, we measured electroencephalography (EEG) activity using a 32 channel BrainAmp system (Brainproducts, Munich, Germany) with Ag/AgCl electrodes (data not analysed here). Participants were seated in an isolated sound-proofed compartment approximately 50 cm in front of a computer screen (Sony Trinitron CPD-E530, 21 inches) with two loudspeakers beside it. During the experiment, participants kept their head in a chin rest and were instructed to sit as still as possible and keep their gaze at the fixation cross on the screen. Before starting the task, participants engaged in a 1-minute-long training round of the FT-RSGT which could be repeated as many times as necessary. After the task, lasting approximately 25 minutes, participants again fill out a PANAS to measure their post-task mood. Then they filled out a custom-made questionnaire regarding the retrospective task motivation (7-point Likert scale ranging from “not very motivated” to “extremely motivated”), and to rate their confidence when responding to the first probe about their task focus (7-point Likert scale ranging from “not at all” to “extremely”). Furthermore, participants provided open-ended feedback about their randomness strategy, comments relating to the task, the content of their MW episodes, and other remarks regarding the experiment in general. If it was the participants first test (T1), they were reminded of their second test (T2) which was always scheduled seven days after their T1 at the same time (either 09:00 or 11:00). If it was their second test (T2), participants received their gift card and were asked whether they wanted a computer-generated summary of their sleep data.

## Statistical analysis

<!--# Bayesian analyses -->

The data analysis was done in `r str_split(R.version$version.string, " \\(")[[1]][[1]]` [@rcoreteam2025]. For our data analyses, we used Bayesian statistics [@wagenmakers2018] with the brms [Bayesian Regression Models using Stan; @bürkner2017] package. For our regression models, we used the default priors, which are non-informative for coefficients corresponding to fixed effects and weakly informative for the intercept and standard-deviation parameters (Student-t prior with 3 df, mean 0 and SD 2.5). The brms package relies on the Stan software [@standevelopmentteam2024] implementing Hamiltonian Mote Carlo sampling. For our models, we ran 6 chains with 6000 iterations each and a warm-up of 3000 iterations. Convergence and stability of the Bayesian sampling was assessed using $\hat{R}$, was ensured to be below 1.05 for all variables [@gelman2019]. We also ensured that the effective sample size (ESS) was greater than 1000 [@bürkner2017]. When reporting Bayesian regression coefficients, we report the posterior mean *b* and its 95% highest-density interval (HDI). In addition, we report the evidence ratio (ER) that the effect is in the (*b*) specified direction. The ER indicates the relative evidence for the effect being in the (*b*) specified ($\text{ER}_\text{dir}$) vs. the opposite direction. For example, a coefficient of 1.2 with an ER of 30 indicates that there is 30 times more evidence that the effect is *positive* (as specified by the *b*) rather than negative. Finally, we report the probability that the effect is in the (*b*) specified direction [$p_\text{dir}$; e.g., $b = -1.2$, $p_\text{dir} = .967$; indicates that there is a 96.7% probability that the effect is *negative* rather than positive; @boayue2020].

<!--# Models - Thought Probes -->

To analyse the thought probes, we implemented Bayesian hierarchical ordinal-probit models [@bürkner2019] to avoid problems with using standard linear models [@boayue2019; @liddell2018]. These types of analyses are routinely used in studies on MW [e.g., @aasen2024; @alexandersen2022; @drevland2024; @boayue2019; @boayue2020; @groot2022] and related fields [e.g., @filmer2019; @turi2019]. For each of the dependent variables MW, MB and S-MW, we included the predictors Time (scaled to 0-1), BV, AE, pre-task positive, pre-task negative and all of their respective interactions with the Sleep (continuous variable), and Sleep independently. For all participants, a random intercept was included in the model, and for the MB and S-MW models, we only considered those reports that were preceded by a MW report (score \>2) .

<!--# Models - Behaviour -->

For analysing task performance, we used two separate Bayesian hierarchical linear regression models for the dependent variables BV and AE. We included the predictors Time (time-on-task, scaled to 0-1), pre-task positive, pre-task negative, their respective interactions with Sleep (NS and PSD), and Sleep independently. We also included a random intercept for each participant. <!--# Models - Mood --> Lastly, we wanted to test whether sleep loss, test time (pre-task vs. post-task) and their interaction, would influence mood scores. To test these relationships, we ran two separate hierarchical Bayesian linear models for each mood score, positive and negative.

# Results

## Baseline Measures

Data from questionnaires collected at T0 are reported in @tbl-baseline-questionnaire. The participants' in our study did not indicate significant sleeping problems and are within the normal ranges of health individuals [in line with @saksvik-lehouillier2020], suggesting that the sample is from a health population. Three participants reported having night shift once per month (one participant did not provide an answer to this question). Most participants (`r sum(general_sleep_renamed$coff_day>0)`) reported to drink coffee, where `r sum(general_sleep_renamed$coff_day<4 & general_sleep_renamed$coff_day>0)` participants reported consuming 3 or fewer cups per day, while `r sum(general_sleep_renamed$coff_day>=4)` reported drinking 4 or more cups a day. As for the consumption of energy drinks, only `r sum(general_sleep_renamed$energy_drink>0)` participants reported consuming them, but maximum one per day.

```{r TBL - baseline questionnaire}
#| label: tbl-baseline-questionnaire
#| tbl-cap: Baseline Sleep, Mood and Beverage Consumption
#| tbl-cap-location: top
#| # Maybe add citations in apa-notes`? `
#| apa-note: ""
#| # META
#| include: TRUE
#| warning: FALSE
#| message: FALSE

outputs[["tbls"]][["baseline_sleep"]]
# citations
# [@krupp1989; jogns1991; @bastien2001; @buysse1989; @torsvall1980; @watson1988; @saksvik-lehouillier2020]
```

## Sleep

Participants self-reported and actigraphy-estimated sleep as well as sleep quality between the two conditions are presented in @tbl-sleep-times-sessions. We found overwhelming evidence that the sleep duration was reduced during the PSD period as compared to the NS period for all. It is noteworthy that, compared to the adjusted sleep, the total self-reported sleep length was longer while it was shorter relative to the actigraphy (see the Supplementary for a comparison). Interestingly, self-reported sleep quality was not affected by PSD (see the Supplementary). Furthermore, although participants were informed not to drink any caffeine-containing beverage before either of the experimental tests, `r demo_pre_test_sum |> filter(name=="no_caffeine" & condition=="control") |> pull(sum)` (`r demo_pre_test_sum |> filter(name=="no_caffeine" & condition=="control") |> pull(na)` missing) participants drank coffee before their NS tests, and `r demo_pre_test_sum |> filter(name=="no_caffeine" & condition=="SD") |> pull(sum)` (`r demo_pre_test_sum |> filter(name=="no_caffeine" & condition=="SD") |> pull(na)` missing) before their PSD tests. Nobody consumed alcohol before any of their test sessions, but `r demo_pre_test_sum |> filter(name=="no_alcohol" & condition == "SD") |> pull(na)` are missing from the NS condition and `r demo_pre_test_sum |> filter(name=="no_alcohol" & condition == "control") |> pull(na)` are missing from the sleep deprivation condition.

```{r TBL - sleep over session and modality}
#| label: tbl-sleep-times-sessions
#| tbl-cap: The Difference Between Sleep and Mood Across Sessions
#| tbl-cap-location: top
#| # META
#| include: true
#| message: false
#| warning: false

outputs[["tbls"]][["sleep_comparison"]]
```

## The Effects of PSD on the Thought Probes

Overall changes in probe responses between sleep conditions can be found in @fig-descriptives-probes.

```{r FIG - descriptive probes}
#| label: fig-descriptives-probes
#| fig-cap: "Changes in Probe Responses Between Conditions and Sample"
#| fig-cap-location: top
#| apa-note: "(a) Each subject is indicated by a triangle/circle and has a gray line connection them. Smaller transparent triangles/circles represent cases that are *excluded* form the reduced sample (see Supplementary). The black squares indicate the grand mean and standard error (SE) for the reduced sample (*n* = 23), while the black cross indicate the grand mean and SE for the whole sample (*n* = 34; i.e., including the smaller transparent cases). (b) Counts of each thought probe and their value between the samples. Note that the overall mean (a) and proportion (b) of each response barely change between sample, suggesing that the exclusion does not significantly impact the proportion of reported thought probe."
#| # META
#| include: true
#| message: false
#| warning: false
#| fig-format: svg
#| fig-height: 8
#| fig-width: 7

outputs[["figs"]][["probes__COMBINED_descriptives"]]
```

<!--# MW results -->

The Bayesian ordered-probit model for the dependent variable MW (@fig-probes-over-time left), indicated overwhelming evidence that MW increased with time after the NS (`r coef_hdi_text(c$mw[,"b_probenum_prop"])`), and increased with more variable responses (`r coef_hdi_text(c$mw[,"b_zlogbv"])`), but no evidence that randomness generation predicted MW (`r coef_hdi_text(c$mw[,"b_zlogapen"])`). Additionally, the model revealed overwhelming evidence that lower pre-task positive mood predicted less MW (`r coef_hdi_text(c$mw[,"b_pre_pos"])`), and weak evidence that pre-task negative mood increased MW (`r coef_hdi_text(c$mw[,"b_pre_neg"])`). After the PSD, the model indicated overwhelming evidence that participants generally reported more MW (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos"])`), but no substantial evidence supported an interaction with time-on-task (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"])`). Furthermore, very strong evidence indicated that sleep deprivation interacted with randomness generation (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:zlogapen"])`; sum: `r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:zlogapen"] + c$mw[,"b_zlogapen"])`), such that MW was predicted by less randomness with increasing amount of sleep deprivation. Strong evidence was found for an interaction between response variability and the PSD (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:zlogbv"])`; sum: (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:zlogbv"] + c$mw[,"b_zlogbv"])`), where increasing sleep deprivation weakened the relationship between BV and MW. No substantial evidence was found for an interaction with pre-task positive (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:pre_pos"])`), but some evidence was found for an interaction with pre-task negative mood (`r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:pre_neg"])`), where being in a more negative mood, after being sleep-deprived, attenuated its predictive effect on MW (sum effect: `r coef_hdi_text(c$mw[,"b_c.Adjusted_Duration.diff.pos:pre_neg"] + c$mw[,"b_pre_neg"])`.

The evidence for some of the coefficients were either attenuated or increased in the reduced model. After the NS, the pre-task positive mood indicated no substantial evidence for reduced MW (`r coef_hdi_text(r$mw[,"b_pre_pos"])`), while the pre-negative mood indicated overwhelming evidence that negative mood reduced MW (`r coef_hdi_text(r$mw[,"b_pre_neg"])`). After the PSD, no substantial evidence favoured a reduction in BV before a MW report (`r coef_hdi_text(r$mw[,"b_sleepdepSD:zlogbv"])`, and some evidence suggested that worse AE tended to precede a MW report (`r coef_hdi_text(r$mw[,"b_sleepdepSD:zlogapen"])`; but the sum effect indicated very strong evidence: `r coef_hdi_text(r$mw[,"b_zlogapen"] + r$mw[,"b_sleepdepSD:zlogapen"])`)

<!--# MB results -->

The corresponding model for the dependent variable MB (@fig-probes-over-time centre) indicated some evidence that MB increased with time-on-task after NS (`r coef_hdi_text(c$mb[,"b_probenum_prop"])`). MB was found to be positively predicted by response variability (`r coef_hdi_text(c$mb[,"b_zlogbv"])`), but no evidence that either randomness generation (`r coef_hdi_text(c$mb[,"b_zlogapen"])`), pre-positive (`r coef_hdi_text(c$mb[,"b_pre_pos"])`) or pre-task negative mood (`r coef_hdi_text(c$mb[,"b_pre_neg"])`) predicted MB. After the PSD, we found overwhelming evidence that participants generally reported more MB (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos"])`). On the other hand, no substantial evidence was found for either an interaction with time-on-task (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"])`), response variability (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos:zlogbv"])`), randomness score (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos:zlogapen"])`), pre-task positive (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos:pre_pos"])`), or pre-task negative mood (`r coef_hdi_text(c$mb[,"b_c.Adjusted_Duration.diff.pos:pre_neg"])`).

In the reduced model, two coefficients had the same direction but indicated stronger evidence. Firstly, the model revealed stronger evidence that MB increased with time-on-task (`r coef_hdi_text(r$mb[,"b_probenum_prop"])`), and that more variable responses tended to occur before a MB report (`r coef_hdi_text(r$mb[,"b_zlogbv"])`). Otherwise, the effect of BV was attenuated to not substantial evidence (`r coef_hdi_text(r$mb[,"b_sleepdepSD:zlogbv"])` ).

<!--# S-MW results  -->

The model for the dependent variable S-MW (@fig-probes-over-time right), after the NS, indicated no evidence that S-MW increased with time-on-task (`r coef_hdi_text(c$smw[,"b_probenum_prop"])`), randomness score (`r coef_hdi_text(c$smw[,"b_zlogapen"])`), or pre-task negative mood (`r coef_hdi_text(c$smw[,"b_pre_neg"])`). Although, some evidence suggested that more variable responses tended to precede S-MW reports (`r coef_hdi_text(c$smw[,"b_zlogbv"])`), very strong evidence suggested that pre-task positive mood decreased S-MW reports in favor of D-MW (`r coef_hdi_text(c$smw[,"b_pre_pos"])`). After the PSD, no substantial evidence was found for a general change in S-MW (PSD: `r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos"])`), nor did it change with time-on-task (`r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"])`), randomness score (`r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:zlogapen"])`), or pre-task negative mood (`r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:pre_neg"])`). However, some evidence suggests an interaction with response variability (`r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:zlogbv"])`; sum effect: `r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:zlogbv"] + c$smw[,"b_zlogbv"])`), where (even) more variable responses tended to precede S-MW reports. Lastly, for the interaction with pre-task positive mood and PSD, we found overwhelming evidence pre-task positive mood no longer predicted S-MW tendency (`r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:pre_pos"])`; sum effect: `r coef_hdi_text(c$smw[,"b_c.Adjusted_Duration.diff.pos:pre_pos"] + c$smw[,"b_pre_pos"])`).

As for the reduced model, only weak evidence indicated that pre-task positive mood reduced the proportion of S-MW reports (`r coef_hdi_text(r$smw[,"b_pre_pos"])`), while stronger evidence was observed for an interaction between PSD and BV (`r coef_hdi_text(r$smw[,"b_sleepdepSD:zlogbv"])`), such that a S-MW tended to preceded by more variable responses.

```{r FIG - changes in probes over time}
#| label: fig-probes-over-time
#| fig-cap: Changes in Probe Responses Over Time
#| fig-cap-location: top
#| apa-note: "Interactions are related to the continuous model."
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-format: svg
#| fig-width: 6.5
#| fig-height: 4

outputs[["figs"]][["probes__PSD_x_time"]]
```

## The Effects of PSD on Task Performance

<!--#  Behavioural variability-->

The Bayesian hierarchical linear regression for the dependent variable BV (@fig-behaviour-over-time left), after the NS, indicated overwhelming evidence that response variability increased with time-on-task (`r coef_hdi_text(c$bv[,"b_probenum_prop"])`), but no substantial evidence that either pre-task positive (`r coef_hdi_text(c$bv[,"b_pre_pos"])`), or pre-task negative mood (`r coef_hdi_text(c$bv[,"b_pre_neg"])`) predicted changes in response variability. Following the PSD, the model revealed overwhelming evidence that participants had generally increased BV (`r coef_hdi_text(c$bv[,"b_c.Adjusted_Duration.diff.pos"])`), and strong evidence that variability further increased over time (PSD $\times$ Trial: `r coef_hdi_text(c$bv[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"])`). Furthermore, overwhelming evidence was found for an interaction with pre-task positive mood (`r coef_hdi_text(c$bv[,"b_c.Adjusted_Duration.diff.pos:pre_pos"])`), where higher positive mood scores, after the PSD, predicted reduced response variability even though the effect appeared to be rather small. Similarly for pre-task negative mood, a negative interaction was found, but only supported by some evidence (`r coef_hdi_text(c$bv[,"b_c.Adjusted_Duration.diff.pos:pre_neg"])`). In the reduced model, we only observed some evidence for the interaction between PSD and pre-task positive (`r coef_hdi_text(r$bv[,"b_sleepdepSD:pre_pos"])`).

<!--# Approximate entropy  -->

The corresponding analysis for the dependent variable AE (@fig-behaviour-over-time right), after the NS, indicated no evidence that randomness score changed with time (`r coef_hdi_text(c$ae[,"b_probenum_prop"])`) or pre-task mood scores (positive: `r coef_hdi_text(c$ae[,"b_pre_pos"])`; negative: `r coef_hdi_text(c$ae[,"b_pre_neg"])`). After the PSD period, no substantial evidence was found for either an initial change in randomness generation (`r coef_hdi_text(c$ae[,"b_c.Adjusted_Duration.diff.pos"])`), and no evidence that PSD interacted with pre-negative mood (`r coef_hdi_text(c$ae[,"b_c.Adjusted_Duration.diff.pos:pre_neg"])`). On the other hand, some evidence was found for an interaction with time as a function of increasing amounts of sleep deprivation (`r coef_hdi_text(c$ae[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"])`; sum: `r coef_hdi_text(c$ae[,"b_c.Adjusted_Duration.diff.pos:probenum_prop"] + c$ae[,"b_probenum_prop"])`), where randomness was reduced over time with the magnitude of PSD. Furthermore, the model also revealed an interaction with pre-task positive mood (`r coef_hdi_text(c$ae[,"b_c.Adjusted_Duration.diff.pos:pre_pos"])`), where randomness score slightly increased with more pre-task positive mood. In the reduced model we observed strong evidence that PSD interacted with Trial (`r coef_hdi_text(r$ae[,"b_sleepdepSD:probenum_prop"])`), such that AE decreased over time more after participants were PSD.

```{r FIG - Changes in Behaviour over time by condition}
#| label: fig-behaviour-over-time
#| fig-cap: Changes in Behaviour Over Time
#| fig-cap-location: top
#| apa-note: ""
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-width: 5
#| fig-height: 3
#| fig-format: svg

outputs[["figs"]][["behaviour__PSD_x_time"]]
```

## The Effects of PSD on Mood

For the dependent variable positive mood, the analysis revealed overwhelming evidence that PSD reduced positive mood (`r coef_hdi_text(mood$p[,"b_c.Adjusted_Duration.diff.pos"])`), and overwhelming evidence that positive mood was reduced after the task (Pre-post: `r coef_hdi_text(mood$p[,"b_prepostpost"])`), but no evidence for an interaction between the PSD and Pre-post task (`r coef_hdi_text(mood$p[,"b_c.Adjusted_Duration.diff.pos:prepostpost"])`). As for the dependent variable negative mood, the analysis revealed strong evidence that PSD increased negative mood (`r coef_hdi_text(mood$n[,"b_c.Adjusted_Duration.diff.pos"])`), and no substantial evidence that the task influenced negative mood (`r coef_hdi_text(mood$n[,"b_prepostpost"])`), but strong evidence suggested that PSD and Pre-post task interacted (`r coef_hdi_text(mood$n[,"b_c.Adjusted_Duration.diff.pos:prepostpost"])`), such that after being the PSD they reported a decrease in negative mood after the task.

```{r FIG - Changes in Behaviour over time by condition}
#| label: fig-psd-on-mood
#| fig-cap: Changes in Mood Over Sleep Loss and Pre-Post Task
#| fig-cap-location: top
#| apa-note: ""
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-width: 5
#| fig-height: 3
#| fig-format: svg

outputs[["figs"]][["Mood__PSD"]]
```

<!--# The interactive effect of PSD on task performance on their associattion to MW -->

```{r FIG - MW interactions from model coefficients}
#| label: fig-interaction-behaviour-on-MW
#| fig-cap: The Interaction Between Behaviour and Sleep Loss on Their Relation to Mind Wandering
#| fig-cap-location: top
#| apa-note: "PSD = Partial sleep deprivation, NS = normal sleep. "
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-width: 5
#| fig-height: 3
#| fig-format: svg

outputs[["figs"]][["MW__PSD_x_behaviour"]]
```

```{r FIG - PSD x mood interacitons - BV + s-mw}
#| label: fig-psd-positive-interactions-bv+smw
#| fig-cap: The Interaction Between Behaviour and Sleep Loss on Their Relation to Mind Wandering
#| fig-cap-location: top
#| apa-note: "PSD = Partial sleep deprivation, NS = normal sleep. "
#| # META
#| include: true
#| warning: false
#| message: false
#| fig-width: 5
#| fig-height: 3
#| fig-format: svg


outputs[["figs"]][["BV+SMW__PSD_x_pre_pos"]]

```


<!--# Sleep quality -->

Sleep quality has been found to correlate positively with time awake, subjective sleep latency and objective sleep efficiency [@elmenhorst2008; @saksvik-lehouillier2020]. Although our participant’s reported sleeping less, they generally did *not* report improved sleep quality. {{{This might suggest that the participants reported their sleep to correspond to the aim of the experiment in an attempt to "behave" according to the experimental demands [i.e., sleeping less, e.g., @mccambridge2014, but see also @mccambridge2012] }}}. Given the variation in sleep reduction between participants, it might be that sleep quality is improved only in those participants who slept even less. However, an exploratory analysis did not find support for this idea: Those who slept even less had similar sleep quality as those who slept more, indicating that the sleep quality in our sample was not influenced by more sleep deprivation. Another possibility could be that the participants phone use, before bed, might have influenced their sleep quality, as previous research has suggested [e.g., @exelmans2019; @hjetland2021; @hysing2015; @lastella2020; @nosetti2021]. Conceivably, if more time spent awake increase sleep quality, but phone use before bed reduce sleep quality, it could be that the effects cancel each other out. Although it should be noted that this is highly speculative as we do not have data on participant's phone use. Thus, future research is needed to confirm whether bedtime phone use impede improvements in sleep quality following sleep deprivation.

<!--# Sleep effects on MW -->

Participants reported more MW after the PSD period compared to after their NS period, in line with previous research. For instance, @poh2016 found that one night of TSD increased MW in a visual search task. Furthermore, a survey study found that poor sleep predicted more MW the following day [@cárdenas-egúsquiza2022b] and a diary study found similar results [@marcusson-clavertz2019]. Building on these findings, a recent systematic review on the relationship between sleep and MW landed on the same conclusion: insufficient sleep, poorer sleep quality, and fragmented sleep all increase disruptive MW [@cárdenas-egúsquiza2022a]. However, it's worth noting that a recent experiment using ambulatory assessment to measure objective and subjective sleep as well as MW throughout the day did not find that changes in sleep predicted changes in MW tendency [@marcusson-clavertz2023]. This contradictory finding might accurately capture the relationship between the natural occurrence of MW and sleep in an everyday situation, but there are many factors, such as hunger, light and temperature that influence our attention [@gabehart2016]. Ultimately, our data demonstrate that MW significantly increased following three nights of reduced sleep.

<!--# Following the PSD, participants generally reported more MW, but indicated no changes over time beyond what is typically found (Figure %%% **SXX** %%%). This increase in MW from insufficient sleep is in line with prior studies. One study found that one night of TSD increased MW for a visual search task [@poh2016]. Furthermore, a survey study suggested that poor sleep predicts more MW the following day [@cárdenas-egúsquiza2022b] and a diary study found similar results [@marcusson-clavertz2019]. A recent meta-review of the literature landed on the same: insufficient sleep {{{ When I note insufficient sleep, I also want to include "worse sleep quality" and "fragmented sleep" (i.e., ineffective sleep) }}} increase disruptive MW [@cárdenas-egúsquiza2022a]. On the other hand, @marcusson-clavertz2023 used an ambulatory assessment method measuring participants objective sleep, subjective sleep and MW throughout a day, and they did *not* find insufficient sleep to predict increases in MW throughout a day. -->

<!--# Sleep effects on MB -->

MB reports were generally low, but higher after the PSD period, as we hypothesized, and some evidence that MB increases with time-on-task. Previous research has found that MB increase with time-on-task during a reading task [@ward2013], but not during a non-cognitive task [@mortaheb2022]. For our task, it might be that the demanding nature of the FT-RSGT could increase the MB reports over time. Interestingly, we found the time-on-task effect after the PSD condition to be slightly smaller compared to the NS condition. One possible explanation is that MB reached a saturation point wherein further sleep deprivation and/or time-on-task cannot further increase MB reports. That could indicate that neuronal processing reaches a point at which it is at balance with the brains' ability to clear out waste [@xie2013] and/or take a break [@vyazovskiy2013]. Indeed, some research suggest that the brain adapts to PSD by downregulating processing capacity to maintain functioning throughout a day [@belenky2003]. Moreover, this interpretation aligns with the resource-depletion account [@jubera-garcía2021a], positing that changes in resource availability could be one cause of MW and MB. More specifically, when resources are drained from certain locations/networks, these locations/networks will need a break to recuperate resources which causes a "down period", where sleep-like slow waves activity can be observed on the scalp during wakefulness [@andrillon2019, @andrillon2021; @jubera-garcía2021b]. Intriguingly, MB appears to be more common when there is a decrease in the activity in the frontal and posterior cortical regions [@andrillon2021; @boulakis2023], suggesting that these regions might be specifically required for conscious awareness of mental content.

<!--# Sleep effects on S-MW -->

Contrary to our hypothesis, the proportion of S-MW was not influenced by the PSD period. Despite the overall increase in MW frequency, participants did not show changes in whether their MW was intentional or not. This finding is intriguing as it suggests that participants engaged in proportionally similar amounts of D-MW even after sleep deprivation, which challenges the prevailing view that sleep loss primarily increases spontaneous attentional lapses [@poh2016; @cárdenas-egúsquiza2022b]. Previous studies have demonstrated that reduced executive functioning - which can be brought about through sleep deprivation [@saksvik-lehouillier2020; @skurvydas2020] - typically leads to more S-MW than D-MW [@seli2016]. This finding could suggest that the moderate reduction in sleep might not have been sufficient to impede executive functioning, which is in line with the finding that AE was not reduced (see below). Alternatively, participants might experience reduced motivation due to being more fatigued, which could have made them more likely to deliberately mind wander. The latter is partially supported by the fact that positive mood attenuated MW reports (see below). Overall, this finding suggests that the relationship between sleep loss and attentional control is more complex than simply increasing S-MW, which necessitates more research to untangle.

<!--# Sleep effects on Behaviour -->

As for the effects on the behaviour, we found the expected pattern for BV but not for AE. In line with our hypothesis, we found BV to increase following the PSD and it also increased with time-on-task. These results align with previous research demonstrating that sleep loss disrupts the stability of cognitive performance [@jackson2013; @doran2001], resulting in greater fluctuations in response timing that further increases with time-on-task [@thomson2014; @zanesco2024a; @zanesco2024b]. However, contrary to our hypothesis, PSD did not impact overall AE performance. This finding is surprising given that executive functions are typically considered vulnerable to sleep deprivation [@alhola2007] and previous research has found reduced randomness generation following TSD [@heuer2005]. It could be that the PSD of 1.7 hours was not sufficient to influence the randomness generation. Alternatively, it could be that the AE measure, in this experiment, might have been unreliable, as we did not find the typical negative relationship between AE and MW [described below; @aasen2024; @drevland2024; @alexandersen2022; @boayue2020]. Moreover, many participants (17) reported not using any particular strategy to create random sequences. However, in an exploratory analysis, comparing those who used no strategy to those who used "any" strategy, revealed that those with "any" strategy did *not* produce more random sequences (see Supplementary). Additionally, one study suggests that the link between AE and MW is not always robust, and can vary depending on contextual factors, like feeling/knowledge of being monitored \[{{{ALEXANDERSEN2024}}}\]. Moreover, this study proposed that performance feedback improves randomness generation, something future research may want to incorporate to improve the reliability of the AE score.

<!--# Sleep effects on Mood -->

The sleep protocol produced clear effects on the participants' affect. They reported significantly reduced positive mood and slightly increased negative mood after the PSD period compared to the NS period. The reduction in positive affect is consistent with previous research demonstrating that moderate sleep restriction can diminish positive mood [@lo2016; @saksvik-lehouillier2020; @palmer2024]. Additionally, we found some evidence for increased negative mood before the task following PSD, though this effect was relatively small and did not persist after the task. We also found a similar decline in positive mood from pre- to post-task, suggesting task engagement reduces mood similarly. This pattern indicates that while PSD lowers the baseline of positive affect, it does not necessarily accelerate mood deterioration during cognitive tasks. These findings align with the emerging view that positive and negative affect may be differentially sensitive to sleep loss, with positive emotions showing greater vulnerability [@palmer2024]. This asymmetrical effect on emotional experience could have important implications for daily functioning, as reduced positive affect might diminish motivation [@seli2015], creativity [@baas2008], and social engagement [@watson1988] even when negative emotions are not markedly increased.

<!--# Discussion on various aspects related to the probes and behaviour -->

In relationship to our tendency to shift our attention, we found several interesting effects. After the NS, MW was typically preceded by more variable responses, in line with previous research [@aasen2024, @boayue2020, @seli2013b], which were slightly attenuated after the PSD. The attenuated variability could stem from the general increase in BV, making BV slightly less predictive of MW. Furthermore, we did not find the expected negative relationship between AE and MW. However, after the PSD period, sequences tended to be less random before a MW episode in line with previous research [@aasen2024, @drevland2024, @alexandersen2022; @groot2022; @boayue2020]. This suggests that sleep deprivation may create conditions where declining executive control becomes more tightly coupled with attentional lapses. The interaction with AE can be interpreted as support for the executive failure view of MW [@mcvay2010], suggesting that sleep loss may impair the executive control mechanisms that typically suppress task-unrelated thoughts. This aligns with neuroimaging literature showing reduced activity in the fronto-parietal control network following sleep deprivation [@yao2023] the same network implicated in MW \[@christoff2009\]. Furthermore, we found a negative association between positive mood and MW that was present regardless of condition. This negative relationship is probably related to motivation aspects that can improve cognitive control [@locke2008] and reduce MW [@seli2015; @unsworth2013]. On the other hand, we did not observe a relationship between negative mood and MW, which is surprising given experimental evidence suggesting a positive relationship between negative mood and MW [@killingsworth2010; @smallwood2009]. Although it should be noted that the relationship between negative mood and MW might be moderated by certain cognitive thinking styles, such as preservative cognition [@ottaviani2013; @ottaviani2015] and guilty/fear-of-failure [@marcusson-clavertz2016]. Thus, only if individuals tend to think in a rigid and inflexible way, does negative mood induce increased MW. At any rate, our sample appeared to have a floor effect for negative mood, and as such, it is challenging to interpret these effects.

For MB, we observed a specific relationship with behavioural performance. The results suggested that a MB report tended to be preceded by more variable responses regardless of condition, compared to a MW episode. The idea that MB constitutes a deeper degree of perceptual coupling [@smallwood2015; @ward2013; @pepin2021] and decreased cognitive processing [@mortaheb2022] is in line with our results, suggesting that MB causes greater instability in performance compared to a MW episode. Indeed, brain activity before a MB episode tends to be highly synchronized [@boulakis2023] which can be disruptive for signal interpretation [@mortaheb2022]. The additive behavioural consequences of MB are particularly worrying for tasks requiring sustained attention, like driving [@poh2016; @albert2018]. Despite the change in BV, we did not observe any changes in AE. This could reflect either a fundamental similarity in how these attentional states affect certain executive functions, or it may relate to the measurement reliability issues with AE discussed earlier. Interestingly, unlike MW, we found no effect of positive or negative mood on MB occurrence, suggesting that MB may represent a distinct attentional state that operates independently of emotional factors. This emotional independence could indicate that MB represents a more basic breakdown in conscious awareness rather than an emotionally-driven shift in attention.

When it comes to MW intention, we found a relationship between mood and attentional control. Specifcally, S-MW indicated a negative relationship with positive mood, suggesting that more positive mood reduced the amount of S-MW. This relationship suggests that positive affect may enhance attentional control, potentially through improved motivation [@seli2015]. When experiencing a more positive mood, participants may have greater cognitive resources available, enabling them to maintain focus on the task or, conversely, to deliberately shift their attention when they choose to do so. This finding aligns with the theoretical framework proposing that D-MW and S-MW reflect different underlying mechanisms [@seli2016]. Positive mood might facilitate a form of cognitive flexibility that, when participants find the task subjectively easier, plausibly through enhanced resources dedicated to the task, they may more often deliberately choose to direct their attention elsewhere. The effect of positive mood on S-MW was not present after the PSD, suggesting that the PSD impeded motivational driven enhancements of cognitive flexibility [@honn2019].

We also observed a relationship between response variability and S-MW after the PSD. Although we found some evidence that a S-MW tended to be related to more variable responses after the NS period, only after the PSD did we find sufficient evidence for this link. One possible explanation is that deliberate shifts of attention only occur when sufficient executive resources are available, such that stable responses could be maintained even though attention is not directly focused on the task. After the PSD, when executive functions are reduced [@palmer2024], spontaneous shifts of attention might be more detrimental for stable responses while D-MW may still require a sufficient pool of resources before being engaged.

Regarding the behavioural outcomes, we observed different effects across our behavioural measures. For BV we observed that pre-task positive mood was negatively related to response variability, but only after sleep deprivation. This protective effect of positive mood might reflect improved motivational engagement with the task [@locke2008; @seli2015]. Interestingly, this relationship was only evident after partial sleep deprivation, unlike the link between positive mood and MW that was apparent in both sleep conditions. One possible explanation is that the generally increased BV following sleep deprivation created conditions where positive mood could have a more pronounced compensatory effect, helping participants with higher positive affect to maintain more stable performance despite being fatigued.

In contrast, AE showed minimal relationship with mood states. The only significant finding for AE was a decrease in randomness generation over time, specifically after sleep deprivation. This time-dependent decline in executive function performance might explain why we observed a relationship between AE and mind wandering only after sleep deprivation - as both measures deteriorated simultaneously later in the task. This pattern suggests that sustained cognitive performance under sleep-deprived conditions may reveal vulnerabilities in executive function that are not apparent under normal sleep conditions. These selective effects highlight the differential sensitivity of various behavioral measures to both mood states and sleep loss, with BV appearing more responsive to emotional factors than randomness generation

## Limitation

This study has several limitations that should be considered. Firstly, one major limitation, is that we did not include a measure of participants’ subjective tiredness before each test. Given that the project started as a student project, this factor might have been so obvious that it managed to slip through. In essence, this means that we cannot be entirely sure whether participants felt more tired after the PSD compared to the NS period. Nevertheless, allowing ourselves to look at the data we do have, we can infer that they probably were more tired after the PSD. We observed that the pre-test mood scores were lower after the PSD compared to the NS, which is a typical effect of sleep restriction [@saksvik-lehouillier2020; @dinges1997; @vandongen2003]. Moreover, the behavioural results suggest that participants performed worse on the FT-RSGT during the PSD test compared to the NS test, suggesting that, something – probably the lack of sleep – influenced their performance.

Secondly, many participants did not sleep sufficiently less. This resulted in many subjects slept below the predetermined threshold of 1.5 hours. As it was undesirable to exclude so many participants, we changed our analytical strategy. We fully acknowledge that this decision can significantly impact the results and the interpretation of the results. Therefore, to be fully transparent, we report the same models in the Supplementary pertaining to the reduced sample, along with a discussion of the changes. Thus, the effects of the mood described in the paper should be interpreted with great care. We, nevertheless, believe that the mood effects described in the model above is more accurate than the one in the reduced sample given the increased sample size.

Thirdly, we did not collect information about potential compensatory strategies during the PSD period. Even though we did inform participants to not take naps during the PSD period, we did not include a measure of whether they took naps. The reason to control for naps is that it has been shown that naps can have an impressive recovering effect on performance in the short-term [i.e., within a couple of hours; @dutheil2021; @leong2022]. However, in our experiment, we find it unlikely that participants took a nap right before the task, as the experiment took place early in the morning. Although, one person reported to the experimenters that they had a nap during the PSD period, but not the test day, due to sleeping way less than the experiment required. It is worth noting that the longer-term recovering effect of naps is less researched. Some evidence suggests that the recovering effects of a nap the day after is rather slim, but nonetheless positive [@vgontzas2007]. Thus, it is plausible that some participants took naps the day(s) before the task and could have recuperated some performances. Future research is advised to include measures of naps in the diary or otherwise.

While the PSD protocol employed here is more ecologically valid than TSD, it is still only a rather short-term intervention. Previous research has investigated longer-term effects of sleep restrictions, such as a week [@dinges1997] and 14 days [@vandongen2003]. However, often, chronic sleep deprivation might be related to 30 minutes less sleep each night for weeks and months at a time [e.g., @sheehan2019; @sivertsen2019]. Such long-term experiments would be more ecologically valid, but more difficult, unpractical and inconvenient for many individuals. Although, some research suggests that there are overlaps between the effects of PSD and TSD [@vandongen2003], much research suggests that these two conditions are dissociable [@belenky2003; @banks2007; @lo2016]. For instance, @belenky2003 suggests that PSD causes adaptive changes in the brain to maintain stable performance by down regulating processing capacity, but this adaptation also comes at the cost of requiring longer recovery, probably because the down regulatory adaptations need to be unwound. Future research may want to investigate how many nights of reduced sleep is sufficient to induce these adaptive changes to make PSD research more efficient, as it would be more convenient for everyone involved.

Some research indicates that sleep variation has a greater influence on performance than length of sleep [@sletten2023; @ness2018]. Thus, it could be that our sample may have had relatively stable sleep times, despite sleeping less, which could be less detrimental for cognitive performance compared to more variable sleep times. The fact that an inconsistent sleep schedule might impact performance more than sleeping less could be related to shifts in the circadian rhythm [@phillips2017] that could impede sleep recovery. Future research may want to investigate how variation in sleep timing impacts performance, especially as sleep variation is a common occurrence [e.g., @phillips2017].

## Summary

We investigated how short-term partial sleep deprivation influences attentional states and cognitive performance. Our sample consisted of healthy adults who experienced partial sleep deprivation, averaging 1.7 hours less sleep each night for three consecutive nights. This moderate sleep reduction led to increased reports of mind wandering and mind blanking, more variable behavioural responses, decreased positive mood, and slightly increased negative mood compared to periods of normal sleep. Notably, PSD established a relationship between response variability and spontaneous MW that was not present during normal sleep, while randomness generation appeared to be unaffected. These findings are concerning when considering the societal trend toward reduced sleep duration [@sheehan2019; @sivertsen2019]. Moreover, our results contribute to the growing literature demonstrating that even modest reductions in sleep significantly impair sustained attention, which is particularly problematic for safety-critical activities such as driving [@rodd2017; @yanko2014].

# References
