---
title: "Supplementary materials for:<br> The Effects of Partial Sleep Deprivation on Mind Wandering, Executive Performance and Affect"
shorttitle: "Supply: PSD on MW, executive performance and affect"
authors:
  - name: Steffen Rygg Aasen
    orcid: 0000-0003-2807-7504
    affiliation: 
      - ref: uit
  - name: Samy Engebø Babiker
    affiliation:
      - ref: uit
  - name: Aurora Marie Kleivik Vangen
    affiliation: 
      - ref: uit 
  - name: Silje Balsnes Haugan
    affiliation: 
      - ref: uit 
  - name: Sanna Aurora Sandell
    affiliation: 
      - ref: uit 
  - name: Ingvild Saksvik-Lehouillier
    orcid: 0000-0002-1953-7364
    affiliation: 
      - ref: ntnu
  - name: Julia W. Y. Kam
    orcid: 0000-0002-2369-2148
    affiliation: 
      - ref: kam
  - name: Matthias Mittner
    orcid: 0000-0003-0205-7353
    affiliation: 
      - ref: uit
  - name: Gábor Csifcsák
    orcid: 0000-0001-7719-7073
    email: gabor.csifcsak@uit.no
    affiliation: 
      - ref: uit 
    corresponding: true
affiliations: 
  - id: uit
    name: "Department of Psychology, UiT The Arctic University of Norway, Norway"
  - id: ntnu
    name: "Department of Psychology, Norwegian University of Science and Technology, Norway"
  - id: kam
    name: "Department of Psychology, Hotchkiss Brain Institute, University of Calgary, Canada"
bibliography: references.bib

# Format
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf: default
  apaquarto-typst: default
floatsintext: true

crossref-tbl-title: "Table S" # Default is "Table" 
crossref-fig-title: "Figure S"# Default is "Figure"

# Code settings
editor: source
execute:
  include: false
  echo: false
fig-cap-location: top
tbl-cap-location: top
cap-location: top
---

```{r Parameters, warning=FALSE, message=FALSE}
relative_path <- "../" # b/c quarto 
source( paste0(relative_path, "lib/quarto_load_project.R") )
quarto_load_project(relative_path, load_data_data = T, load_data_scripts = T, load_caches = F)
```

# Supplementary Materials

## Sleep

Participant 31 did not self-report their normal sleep before the NS test and received special treatment. We used the latest nights of sleep as a proxy for their 3 nights of NS before the NS test.

<!-- Overall sleep effect -->

Participants demonstrated significantly reduced sleep duration during the PSD period compared to the NS period, as confirmed by both the self-reported, actigraphy and the adjusted sleep estimates. However, considerable inter-individual variability was observed in the magnitude of sleep restriction achieved. Previous research demonstrates that sensitivity to sleep loss varies between, but remains consistent within individuals [@hudson2020; @rupp2012; @vandongen2004; @dennis2017]. This suggests that participants may have naturally modulated their sleep restriction according to their own sensitivity to sleep loss. However, the variability in sleep loss could have occurred because of ambiguous sleep restriction instructions. Sleep restriction protocols typically employ either a fixed time-in-bed [e.g., @hudson2020; @vandongen2004] or predetermined bedtimes [@saksvik-lehouillier2020]. In contrast, our protocol required participants to estimate their own sleep restriction, potentially allowing for less stringent implementation among some individuals. Therefore, the inter-individual variability in sleep loss likely reflects both individual differences in sleep loss and methodological limitations of the protocol.

<!--# We have some discrepancies, explain how these might arise -->

The self-report and actigraphy did not always provide similar sleep/wake times. The reason for the discrepancies likely stems from two sources: First, participants might have adjusted their reports to meet perceived study expectations (e.g., reporting later bedtimes or earlier wake times than actually experienced). Second, actigraphy has inherent limitations in accurately detecting sleep/wake transitions. For example, actigraphy might misinterpret motionless pre-sleep activities as sleep - such as using a mobile phone in bed, a common phenomenon in modern society [@gradisar2013; @hjetland2021; @hysing2015]. Similarly, it might fail to register actual wake times when participants remain in bed after waking. Regardless of these measurement challenges, our data showed that participants generally slept less during the PSD period compared to the NS period, as indicated by the self-report, actigraphy and most importantly, the adjusted sleep estimates that takes both measures into consideration.

### Shiney

In the Shiny application [@shiney2025; see the application at <https://github.com/cogneuro-uit/mw-psd-eeg--2021--behaviour/tree/master/shiny-sleep-adjustments>], time was visualized on the x-axis, while the actigraphy recorded activity (raw data) and light (log-transformed) were visualized on separate y-axes. Coloured marks highlighted the self-reported and actigraphy sleep onset and wake times. The rest interval was indicated by shaded areas ranging from the sleep attempt to the sleep onset, and the wake time to the rise time for the self-report and actigraphy. The researchers could either adjust two sliders to adjust the time (hours and minutes), providing a third coloured marker, or use two separate buttons to apply the exact values of the self-reported or actigraphy as the onset/wake time. Around each marker, a transparently coloured area of (+-) 15 minutes was presented.

### Adjusted Sleep Estimate

@fig-onset-wake-diffs indicates the differences between self-reported and actigraphy onset/wake times.

```{r}
#| label: fig-onset-wake-diffs
#| fig-cap: Differences Between Self-Reported and Actigraphy Sleep Onset and Wake
#| fig-cap-location: top
#| fig-format: svg
#| fig-width: 4.5
#| fig-height: 8
#| fig-dpi: 300
#| apa-note: "(a) Indicate the overall difference between the self-report and actigraphy estimated sleep onset/wake across condition. (b) Indicate the individual relationship between self-reported and actigraphy across conditions. NS = normal sleep, PSD = partial sleep deprivation."
#| ##  META  ##
#| include: true
#| warning: false
#| message: false

figs[["onset-wake_difference"]]
```

@fig-SD-between-modalities indicates the estimated sleep loss during the partial sleep deprivation (PSD) compared to the normal sleep (NS). Note some great discrepancies between red (self-reported) and blue (actigraphy) estimates.

```{r}
#| label: fig-SD-between-modalities
#| fig-cap: Difference in Sleep Duration Between Sources 
#| fig-cap-location: top
#| fig-format: svg
#| fig-width: 9.5
#| fig-dpi: 300
#| apa-note: "Sleep deprivation is estimated as the sleep lost during the partial sleep deprivation (PSD) period relative to the normal sleep (NS) period. Sleep duration is averaged for the three nights preceding each test, and the difference is calculated as PSD - NS. The black solid line indicates the original acceptance threshold, while the red dashed line indicates the adjusted sleep threshold (the criterion used for the reduced sample). Cases above the 1.5 threshold are highlighted by their associated participant code."
#| ##  META  ##
#| include: true
#| warning: false
#| message: false

figs[["PSD__report_type"]]
```

@fig-sleeptimes-across-source illustrates how self-reported, actigraphy and adjusted sleep variables related to mind wandering (MW).

```{r}
#| label: fig-sleeptimes-across-source
#| fig-cap-location: top
#| fig-cap: Relationship Between Sleep Loss and Mind Wandering Across Sleep Estimates
#| fig-height: 5
#| fig-dpi: 600
#| apa-note: "Note. Average sleep deprivation is calculated based on the difference from the normal sleep (NS) period, such that 0 indicates each subject’s NS sleep time, and the sleep deprivation indicates the (average) differences from their NS average. The solid lines indicate the overall relationship for the three different measure types. PSD = partial sleep deprivation."
#| ##  Meta  ##
#| include: true 
#| warning: false 
#| message: false

figs[["sleep_variables_mw"]]
```

@fig-sleeptimes-across-source-summarised indicates the sleep estimates across sleep type and a connecting line between each participant.

```{r}
#| label: fig-sleeptimes-across-source-summarised
#| fig-cap-location: top
#| fig-cap: Relationship Between Sleep Loss and Mind Wandering Across Sleep Estimates
#| fig-height: 5
#| fig-dpi: 600
#| apa-note: "Note. Average sleep deprivation is calculated based on the difference from the normal sleep (NS) period, such that 0 indicates each subject’s NS sleep time, and the sleep deprivation indicates the (average) differences from their NS average. The solid lines indicate the overall relationship for the three different measure types. PSD = partial sleep deprivation."
#| ##  Meta  ##
#| include: true 
#| warning: false 
#| message: false

figs[["PSD__report_type_summarised"]]
```

### Test of Differences Between Sleep Estimates Between Self-Reports and Actigraphy

We explored whether there was any significant difference between the modalities (self-report and actigraphy). @tbl-diff-sr-ag shows the result of a simple paired sample Bayesian t-test, indicating that sleep duration but not sleep onset and wake time differed between modalities.

```{r}
#| label: tbl-diff-sr-ag
#| tbl-cap: The Difference in Sleep Estimates Between Self-Report and Actigraphy
#| tbl-cap-location: top
#| ##  META  ##
#| include: true
#| warning: false
#| message: false

tbls[["test_sleep_vars_across_modality"]]
```

<!-- ```{r} -->

<!-- #| label: tbl-test-diff-sr-ag -->

<!-- #| tbl-cap: "The Difference in Sleep Estimates Between Self-Report and Actigraphy" -->

<!-- #| tbl-cap-location: top -->

<!-- #| ##  META  ## -->

<!-- #| include: true -->

<!-- #| warning: false -->

<!-- #| message: false -->

<!-- tbls[["test_sleep_wake_delay"]] -->

<!-- ``` -->

## Probes Descriptives

@fig-desc-additional-probe-comparison indicates the total raw count of probe values across probe type, condition, and data treatment. @tbl-probe-raw-count illustrates additional comparisons between the probe response options across conditions and sample choice. Note that, regardless of sample size choice, the proportions of responses remain comparable, suggesting that the exclusion of individuals did not change the response distribution meaningfully.

```{r}
#| label: tbl-probe-raw-count
#| tbl-cap: "Total Raw Count of Probe Values Across Probe Type, Condition, and Data Treatment"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["probe_descriptives_wide"]]
```

```{r}
#| label: fig-desc-additional-probe-comparison
#| fig-cap-location: top
#| fig-cap: "Relative Frequency and Count of Probe Responses Over Probe Type, Condition, and Data Treatment"
#| fig-height: 8.5
#| fig-width: 8
#| fig-dpi: 300
#| apa-note: "Note. a) Shows the relative frequency of each probe response across probe question, condition, and sample treatment. Similarly, b) shows the raw count across each participant (partially transparent) with an overall grand average (solid colour)."
#| ##  Meta  ##
#| include: true 
#| warning: false 
#| message: false

figs[["desc_probes_additional_comparisons"]]
```

## Dose-Response Models

@tbl-dr-probes-models indicates the coefficients for the continuously treated PSD on the various MW types, and @tbl-dr-behavioural-models indicates the coefficients for the continuously treated PSD on the behaviours. Bold rows indicate a difference between the DR and SC models (see below).

```{r}
#| label: tbl-dr-probes-models
#| tbl-cap: "Dose-Response Bayesian Model for Mind Wandering, Mind Blanking and Spontaneous Mind Wandering"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["cont_probit_model"]]
```

```{r}
#| label: tbl-dr-behavioural-models
#| tbl-cap: "Dose-Response Bayesian Model for Mind Wandering, Mind Blanking and Spontaneous Mind Wandering"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["dich_behav_models"]]
```

## Strict Compliance (SC) Models

### Participants

In the strict compliance (SC) models, we treated the partial sleep deprivation as a dichotomous variable. Using our pre-defined criteria for sleep loss, `r subj$red$exc` participants had to be excluded. This left the SC models with a total sample size of `r subj$red$num` participants with an average age of `r subj$red$age` (*SD* = `r subj$red$age_sd`, range = `r subj$red$age_min`-`r subj$red$age_max`, females = `r subj$red$gender_f`).

## Bayesian Statistical Models

### Thought Probes

The Bayesian hierarchical ordered-probit models of the thought probes for the SC models are presented in @tbl-dr-probes-models. 

```{r}
#| label: tbl-dr-probes-models
#| tbl-cap: "Dose-Response Bayesian Model for Mind Wandering, Mind Blanking and Spontaneous Mind Wandering"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["dich_probit_model"]]
```

### Behaviour

The Bayesian hierarchical linear model of the behaviour for the SC models are presented in @tbl-sc-behavioural-models.

```{r}
#| label: tbl-sc-behavioural-models
#| tbl-cap: "Dose-Response Bayesian Model for Mind Wandering, Mind Blanking and Spontaneous Mind Wandering"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["dich_behav_models"]]
```


::: landscape

## Affect

### Pre-Task Affect

```{r}
#| label: tbl-psd-on-pre-task-affect
#| tbl-cap: "Partial Sleep Deprivation on pre-Task Positive and Negative Affect"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["psd_on_pre-task_affect"]]
```

### Affect Change Across the Task

```{r}
#| label: tbl-psd-mw-on-pre-task-affect
#| tbl-cap: "The Effect of Partial Sleep Deprivation and Mind Wandering on the Change in Affect Across the Task"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["affect-change__psd+mw"]]
```


:::

## Collider bias

Excluding participants based on an arbitrary threshold can be concerning as such exclusion could be caused by another underlying factor, like motivation. To ensure that participants included in the SC models did not systematically differ from those that were excluded we, we used both Welch t-tests and Bayesian t-tests to test the difference across a various metrics. Specifically, we tested the groups across our probe responses (MW, MB, S-MW), task performance indices (BV, AE), and self-reported questionnaires (pre-positive mood, pre-negative mood, fatigue, sleepiness, insomnia, sleep quality, diurnal preference, alcohol consumption). Our results (@tbl-compliance_test_excluded_vs_included) indicated that there were no baseline differences between the groups, suggesting that the two groups did not systematically differ across various metrics other than their ability to achieve the specified sleep criteria.


```{r}
#| label: tbl-compliance_test_excluded_vs_included
#| tbl-cap: "The Effect of Order on Thought Probes"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["compliance_test_excluded_vs_included"]] |> tab_header(title = "")
```

## Order effects

To test whether there was an effect of order, we ran 3 Bayesian hierarchical ordered-probit models of the outcomes MW, MB, S-MW, and 2 Bayesian hierarchical linear models for the outcomes BV and AE with the predictors PSD and session (S1 vs S2), including random intercepts. Generally, the results did not indicate any order effects, except for MB, which generally showed reduced MB during S2 (b = -0.34, ERdir = 40.1, pdir = .976). However, that effect was qualified by the S2 x PSD interaction, such that whilst MB was generally reduced during S2, this reduction diminished with increasing PSD (b = 0.34, ERdir = 114, pdir = .991). Similarly, there appears to be a training effect for AE, which showed a general increase in the second session, but did not interact with PSD. Crucially, the PSD effect on MW tendency remains even when we control for order. Results are displayed in @tbl-order_effects_on_probes and @tbl-order_effects_on_behav.

```{r}
#| label: tbl-order_effects_on_probes
#| tbl-cap: "The Effect of Order on Thought Probes"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["order_effects_on_probes"]] |> tab_header(title = "")
```

```{r}
#| label: tbl-order_effects_on_behaviour
#| tbl-cap: "The Effect of Order on Behaviour"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["order_effects_on_behav"]] |> tab_header(title = "")
```

## Test Time

We tested whether test time (09:00 vs. 11:00) had an effect on the thought probes and behaviour. To achieve this, we implemented 5 Bayesian hierarchical linear models with outcomes MW, MB, S-MW, BV, and AE using the predictors PSD and test time (09:00 vs. 11:00). For the chronotype model, we did a simple Bayesian t-test of the average chronotype split by the test time group. For the chronotype model, those who were tested earlier tended to have an earlier chronotype ($M$ = `r effect_of_test_time$m_9`, $SD$ = `r effect_of_test_time$sd_9`) while those who were tested later tended to have a later chronotype ($M$ = `r effect_of_test_time$m_11`, $SD$ = `r effect_of_test_time$sd_11`). However, we did not observe any evidence ($BF_{10}$ = `r effect_of_test_time$bf`) that this difference ($M_\text{diff} $*M*~diff~ = 0.29) was significant. For the hierarchical linear models, the results did not find any effect of test time on the outcomes (none of the HDIs of the coefficients excluded zero) while maintaining the PSD effect (@tbl-test_time_on_probes and @tbl-test_time_on_behaviour).

```{r}
#| label: tbl-test_time_on_probes
#| tbl-cap: "Test Time Effects on the Thought Probes"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["test_time_on_probes"]] |> tab_header(title = "")
```

```{r}
#| label: tbl-test_time_on_behaviour
#| tbl-cap: "Test Time Effects on Behaviour"
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["test_time_on_behav"]] |> tab_header(title = "")
```

{{< pagebreak >}}

# References
