---
title: "Supplementary materials for:<br> The Effects of Partial Sleep Deprivation on Mind Wandering, Executive Performance and Affect"
shorttitle: "Supply: PSD on MW, executive performance and affect"
authors:
  - name: Steffen Rygg Aasen
    orcid: 0000-0003-2807-7504
    affiliation: 
      - ref: uit
    roles: 
      - writing
      - data curation
      - investigation
      - formal analysis
      - resources
      - visualization
  - name: Samy Engebø Babiker
    affiliation:
      - ref: previously
      - ref: uit
    roles: 
      - data curation
      - investigation
      - resources
      - editing
  - name: Aurora Marie Kleivik Vangen
    affiliation: 
      - ref: uit 
    roles: 
      - data curation
      - investigation
      - resources
  - name: Silje Balsnes Haugan
    affiliation: 
      - ref: previously
      - ref: uit 
    roles: 
      - data curation
      - investigation
      - resources
      - editing
  - name: Sanna Aurora Sandell
    affiliation: 
      - ref: previously
      - ref: uit 
    roles: 
      - data curation
      - investigation
      - resources
  - name: Gábor Csifcsák
    orcid: 0000-0001-7719-7073
    affiliation: 
      - ref: uit 
    roles: 
      - conceptualization
      - methodology
      - supervision
      - validation
      - resources
      - editing
  - name: Ingvild Saksvik-Lehouillier
    orcid: 0000-0002-1953-7364
    affiliation: 
      - ref: ntnu
    roles: 
      - editing
      - resources
      - methodology
  - name: Julia W. Y. Kam
    orcid: 0000-0002-2369-2148
    affiliation: 
      - ref: kam
    roles:
      - editing
  - name: Matthias Mittner
    orcid: 0000-0003-0205-7353
    email: matthias.mittner@uit.no
    affiliation: 
      - ref: uit
    roles: 
      - conceptualization
      - funding acquisition
      - project administration
      - methodology
      - software
      - resources
      - supervision
      - validation
      - visualization
      - editing
    corresponding: true
affiliations: 
  - id: uit
    name: "Department of Psychology, UiT The Arctic University of Norway, Norway"
  - id: ntnu
    name: "Department of Psychology, Norwegian University of Science and Technology, Norway"
  - id: kam
    name: "Department of Psychology, Hotchkiss Brain Institute, University of Calgary, Canada"
  - id: previously
    name: "Previously associated with the"
keywords: [attention, sleep, mind wandering, partial sleep deprivation, executive functions, affect, mood]
author-note:
  disclosures:
    conflict-of-interest: "The author has no conflict of interest to declare. The authors wish to declare that they have used large language models (LLM), such as Claude, ChatGPT, SciSpace, Perplexity, to aid the various steps in the production of this manuscript (e.g., writing, literature search, analysis and coding). Any added material was always fact-checked and verified before being added to the manuscript."
    data-sharing: "Data from this study can be accessed at the Open Science Foundation (OSF): https://osf.io/xq6wr/."
    financial-support: ""
    gratitude: "We would like to extend our gratitude to Ragnhild Nicolaisen Drevland for helping in the data collection, as well as to our participants who slept less (for Science!)."
bibliography: references.bib

# Format
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf: default
  apaquarto-typst: default
floatsintext: true

crossref:
  fig-prefix: "Figure S"     # Default is "Figure"
  tbl-prefix: "Table S"      # Default is "Table"

# Code settings
editor: source
execute:
  include: false
  echo: false
fig-cap-location: top
tbl-cap-location: top
cap-location: top
---

```{r Parameters, warning=FALSE, message=FALSE}
relative_path <- "../" # b/c quarto 
source( paste0(relative_path, "lib/quarto_load_project.R") )
quarto_load_project(relative_path, load_data_data = T)
```

# Supplementary Materials

## Sleep 

Participant 31 did not self-report their normal sleep before the NS test and received special treatment. We used the latest nights of sleep as a proxy for their 3 nights of NS before the NS test. 

<!-- Overall sleep effect -->
Participants demonstrated significantly reduced sleep duration during the PSD period compared to the NS period, as confirmed by both the self-reported, actigraphy and the adjusted sleep estimates. However, considerable inter-individual variability was observed in the magnitude of sleep restriction achieved. Previous research demonstrates that sensitivity to sleep loss varies between, but remains consistent within individuals [@hudson2020; @rupp2012; @vandongen2004; @dennis2017]. This suggests that participants may have naturally modulated their sleep restriction according to their own sensitivity to sleep loss. However, the variability in sleep loss could have occurred because of ambiguous sleep restriction instructions. Sleep restriction protocols typically employ either a fixed time-in-bed [e.g., @hudson2020; @vandongen2004] or predetermined bedtimes [@saksvik-lehouillier2020]. In contrast, our protocol required participants to estimate their own sleep restriction, potentially allowing for less stringent implementation among some individuals. Therefore, the inter-individual variability in sleep loss likely reflects both individual differences in sleep loss and methodological limitations of the protocol. 

<!--# We have some discrepancies, explain how these might arise -->

The self-report and actigraphy did not always provide similar sleep/wake times. The reason for the discrepancies likely stems from two sources: First, participants might have adjusted their reports to meet perceived study expectations (e.g., reporting later bedtimes or earlier wake times than actually experienced). Second, actigraphy has inherent limitations in accurately detecting sleep/wake transitions. For example, actigraphy might misinterpret motionless pre-sleep activities as sleep - such as using a mobile phone in bed, a common phenomenon in modern society [@gradisar2013; @hjetland2021; @hysing2015]. Similarly, it might fail to register actual wake times when participants remain in bed after waking. Regardless of these measurement challenges, our data showed that participants generally slept less during the PSD period compared to the NS period, as indicated by the self-report, actigraphy and most importantly, the adjusted sleep estimates that takes both measures into consideration.

### Shiney

In the Shiny application [@shiney2025; see the application at <https://github.com/cogneuro-uit/mw-psd-eeg--2021--behaviour/tree/master/shiny-sleep-adjustments>], time was visualized on the x-axis, while the actigraphy recorded activity (raw data) and light (log-transformed) were visualized on separate y-axis. Coloured marks highlighted the self-reported and actigraphy sleep onset and wake times. The rest interval was indicated by shaded areas ranging from the sleep attempt to the sleep onset, and the wake time to the rise time for the self-report and actigraphy. The researchers could either adjust two sliders to adjust the time (hours and minutes), providing a third coloured marker, or use two separate buttons to apply the exact values of the self-reported or actigraphy as the onset/wake time. Around each marker, a transparently coloured area of (+-) 15 minute was presented.


### Adjusted Sleep Estimate 

@fig-onset-wake-diffs indicates the differences between self-reported and actigraphy onset/wake times. 

```{r}
#| label: fig-onset-wake-diffs
#| fig-cap: Differences Between Self-Reported and Actigraphy Sleep Onset and Wake
#| fig-cap-location: top
#| fig-format: svg
#| fig-width: 4.5
#| fig-height: 8
#| fig-dpi: 300
#| apa-note: "(a) Indicate the overall difference between the self-report and actigraphy estimated sleep onset/wake across condition. (b) Indicate the individual relationship between self-reported and actigraphy across conditions. NS = normal sleep, PSD = partial sleep deprivation."
#| ##  META  ##
#| include: true
#| warning: false
#| message: false

figs[["onset-wake_difference"]]
```

@fig-SD-between-modalities indicates the estimated sleep loss during the partial sleep deprivation (PSD) compared to the normal sleep (NS). Note some great discrepancies between red (self-reported) and blue (actigraphy) estimates. 

```{r}
#| label: fig-SD-between-modalities
#| fig-cap: Difference in Sleep Duration Between Sources 
#| fig-cap-location: top
#| fig-format: svg
#| fig-width: 9.5
#| fig-dpi: 300
#| apa-note: "Sleep deprivation compared to the normal sleep (NS). Sleep duration is averaged for the three nights preceding each test, and the difference is calculated as partial sleep deprivation (PSD) - NS. The black solid line indicates the original acceptance threshold, while the red dashed line indicated the adjusted sleep threshold (the criterion used for the reduced sample). Cases above the 1.5 threshold are highlighted by their associated participant code."
#| ##  META  ##
#| include: true
#| warning: false
#| message: false

figs[["PSD__report_type"]]
```

@fig-sleeptimes-across-source illustrates how self-reported, actigraphy and adjusted sleep variables related to mind wandering (MW). 

```{r}
#| label: fig-sleeptimes-across-source
#| fig-cap-location: top
#| fig-cap: Relationship Between Sleep Loss and Mind Wandering Across Sleep Estimates
#| fig-height: 5
#| fig-dpi: 300
#| apa-note: "NS = normal sleep, PSD = partial sleep deprivation"
#| ##  Meta  ##
#| include: true 
#| warning: false 
#| message: false

figs[["sleep_variables_mw"]]
```

### Test of Differences Between Sleep Estimates Between Modality

We explored whether there was any significant difference between the modalities (self-report and actigraphy). @tbl-diff-sr-ag shows the result of a simple paired sample Bayesian t-test, indicating that sleep duration but not sleep onset and wake time differed between modalities.

```{r}
#| label: tbl-diff-sr-ag
#| tbl-cap: The Difference in Sleep Estimates Between Self-Report and Actigraphy
#| tbl-cap-location: top
#| ##  META  ##
#| include: true
#| warning: false
#| message: false

tbls[["test_sleep_vars_across_modality"]]
```

### Sleep and Wake Delay

@fig-delay-in-sleep-and-wake indicates the self-reported delay from sleep attempt to sleep onset and from wake time to rise time. @tbl-test-diff-sr-ag indicate the test statistics, using a simple Bayesian paired-sample t-test, for the sleep delay

```{r}
#| label: fig-delay-in-sleep-and-wake
#| fig-cap: Self-Reported Delay in Sleep Onset and Wake Time
#| fig-cap-location: top
#| fig-format: svg
#| fig-width: 6
#| fig-dpi: 300
#| apa-note: "PSD = partial sleep deprivation, NS = normal sleep"
#| ## Settings
#| include: true
#| message: false
#| warning: false

figs[["delay_in_wake-onset"]]
```


```{r}
#| label: tbl-test-diff-sr-ag
#| tbl-cap: The Difference in Sleep Estimates Between Self-Report and Actigraphy
#| tbl-cap-location: top
#| ##  META  ##
#| include: true
#| warning: false
#| message: false

tbls[["test_sleep_wake_delay"]]
```


### The Relationship Between Sleep Quality and Reduced Sleep

To test whether sleep quality improved with greater PSD, we used a simple Bayesian linear model with sleep quality as the outcome and sleep loss as the predictor. Results are shown presented in @tbl-sleep-quality and illustrated in @fig-sleep-quality-over-sleep-loss, indicating no relationship between sleep loss (hour of reduced sleep) and sleep quality. 

<!--# Sleep quality -->

Sleep quality has previously been reported to be positively correlated with time spent awake and objective sleep quality, and negatively correlated with sleep latency [@elmenhorst2008; @saksvik-lehouillier2020]. Although our participants reported sleeping less during the PSD period, they did not report better sleep quality. Given the variation in sleep reduction between participants, sleep quality might have improved only in participants who slept the least. However, an exploratory analysis (see the Supplementary) did not find support for this idea: Those who slept the least had similar sleep quality as those who slept more, indicating that the PSD did not influence sleep quality in our sample. One alternative explanation is that participants used their phone before falling asleep, which could have reduced their sleep quality [e.g., @exelmans2019; @hjetland2021; @hysing2015; @lastella2020; @nosetti2021], possibly through blue-light [@chang2015]. However, this alternative explanation is speculative as we do not have data on participant's phone use, and future research is needed to confirm whether bedtime phone use impedes improvements in sleep quality following sleep deprivation.

```{r}
#| label: tbl-sleep-quality
#| tbl-cap: The Relationship Between Sleep Loss and Sleep Quality
#| tbl-cap-location: top
#| ##  META  ##
#| include: true
#| warning: false
#| message: false

tbls[["sleep_quality+sleep"]] 
```


```{r}
#| label: fig-sleep-quality-over-sleep-loss
#| fig-cap: The Effect of the Partial Sleep Deprivation on Sleep Quality
#| fig-cap-location: top
#| fig-format: svg
#| fig-width: 5.5
#| fig-dpi: 300
#| apa-note: "PSD = partial sleep deprivation, NS = normal sleep"
#| ## META ## 
#| include: true
#| message: false
#| warning: false

PSD_sleep_quality |>
  ggplot(aes(sleep, sleep_quality_SD, col = name)) +
  # facet_wrap(~name) +
  geom_point(alpha = .5) + 
  geom_smooth(method = "lm", alpha = .07)  +
  labs(x = "Sleep deprivation", y = "Sleep quality", col = "Modality") +
  theme(legend.position = "top")
```


## Dichotomous model

### Participants

@tbl-dichotomous-demo indicates the dichotomous sample demographics. 

```{r}
#| label: tbl-dichotomous-demo
#| tbl-cap: Dichotomous Sample Demographics
#| tbl-cap-location: top
#| ##  META  ##
#| include: true
#| warning: false
#| message: false

demographics_reduced |> 
  mutate(Age = as.numeric(Age)) |>
  summarise(
    sum = length(Age)
    , female = sum(Gender == 1)
    , male = sum(Gender == 2)
    , missing = sum(is.na(Age))
    , min = min(Age, na.rm = T)
    , mean = mean(Age, na.rm = T)
    , sd = sd(Age, na.rm = T)
    , max = max(Age, na.rm = T)
    , skewness = moments::skewness(Age, na.rm = T)
    , kurtosis = moments::kurtosis(Age, na.rm = T)
  ) |>
  mutate(across(c(mean, sd, skewness, kurtosis), ~fmt_APA_numbers(.x, .chr=T))) |>
  gt() |>
  tab_spanner("Age", c(min, mean, sd, max, skewness, kurtosis)) |>
  tab_spanner("Number", c(sum, female, male, missing)) |>
  cols_label(
    sum = "Total", female = "Female", male = "Male", missing = "Missing",
    min = "Min", mean = md("*M*"), sd = md("*SD*"), max = "Max",
    skewness = "Skew", kurtosis ="Kurt"
  )
```


### Bayesian Statistical Models

#### Thought Probes

The Bayesian hierarchical ordered-probit models of the thought probes for the dichotomous variable treatment is presented in @tbl-dich-probes. Changes in the probe responses over time and conditions are illustrated in @fig-dich-probes-over-time.

```{r}
#| label: tbl-dich-probes
#| tbl-cap: Dichotomous Bayesian Models for Mind Wandering, Mind Blanking and Spontaneous Mind Wandering
#| tbl-cap-location: top
#| ## META ##
#| include: true
#| warning: false
#| message: false

tbls[["dich_probit_model"]]
```

```{r}
#| label: fig-dich-probes-over-time
#| fig-cap: "Thought Probes Over Time Between Conditions"
#| fig-cap-location: top
#| fig-width: 8
#| fig-height: 4.5
#| fig-dpi: 300
#| ## META ##
#| include: true
#| warning: false
#| message: false

figs[["dich_probes_over_time"]]
```

#### Behaviour

The Bayesian hierarchical linear model of the behaviour for the dichotomous variable treatment is in @tbl-dich-behav-models. Changes in behaviour over time and conditions are illustrated in @fig-dich-behav-over-time.

```{r}
#| label: tbl-dich-behav-models
#| tbl-cap: Bayesian Behavioural Models for The Reduced Sample
#| tbl-cap-location: top
#| # settings
#| include: true
#| warning: false
#| message: false

tbls[["dich_behav_models"]]
```

```{r}
#| label: fig-dich-behav-over-time
#| fig-cap: "Behaviour Over Time Between Conditions"
#| fig-cap-location: top
#| fig-width: 8
#| fig-height: 4
#| fig-dpi: 300
#| # settings
#| include: true
#| warning: false
#| message: false

figs[["dich_behav_over_time"]] 
```

